A descending inhibitory mechanism of nociception mediated by an evolutionarily conserved neuropeptide system in Drosophila

Version 1

Abstract

Abstract

Nociception is a neural process that animals have developed to avoid potentially tissue-damaging stimuli. While nociception is triggered in the peripheral nervous system, its modulation by the central nervous system is a critical process in mammals, whose dysfunction has been extensively implicated in chronic pain pathogenesis. The peripheral mechanisms of nociception are largely conserved across the animal kingdom. However, it is unclear whether the brain-mediated modulation is also conserved in non-mammalian species. Here, we show that Drosophila has a descending inhibitory mechanism of nociception from the brain, mediated by the neuropeptide Drosulfakinin (DSK), a homolog of cholecystokinin (CCK) that plays an important role in the descending control of nociception in mammals. We found that mutants lacking dsk or its receptors are hypersensitive to noxious heat. Through a combination of genetic, behavioral, histological, and Ca2+ imaging analyses, we subsequently revealed neurons involved in DSK-mediated nociceptive regulation at a single-cell resolution and identified a DSKergic descending neuronal pathway that inhibits nociception. This study provides the first evidence for a descending modulatory mechanism of nociception from the brain in a non-mammalian species that is mediated by the evolutionarily conserved CCK system, raising the possibility that the descending inhibition is an ancient mechanism to regulate nociception.

Introduction

Minimizing tissue damage is a fundamental task for all animals to increase their chance of survival. Thus, elucidating the principles of nociception, the neural process detecting and encoding potentially tissue-damaging stimuli, is critical to understanding the molecular and neural mechanisms implementing adaptive behaviors and their evolution. Nociceptors are sensory neurons specialized to detect harmful stimuli, whose activation triggers downstream nociceptive circuits and nocifensive responses (1). Since the activities of nociceptors and downstream nociceptive circuits are tightly linked to pain perception in humans, unveiling the mechanisms of nociception is also crucial to a better understanding of human pain mechanisms (2, 3).

Descending inhibition has been suggested to be a pivotal mechanism in the modulation of nociception and pain in mammals. Since the discovery that electrical stimulations of parts of the midbrain in rats enabled surgical operations without anesthetics (4), mammalian descending nociceptive pathways have been implicated in various analgesic phenomena/treatments and the development of chronic pain states, suggesting their critical role in modulating nociception and pain (5, 6). However, brain-mediated modulatory mechanisms of nociception such as descending inhibition have currently been identified only in mammals, despite a high degree of commonality across species in the peripheral nociceptive mechanisms (7, 8). Therefore, it is unknown whether the descending modulation is a de novo mechanism typical of the highly developed mammalian central nervous system (CNS) or a conserved control also present in simpler animals.

The descending nociceptive-modulatory systems in mammals have been revealed to involve various neurochemical pathways (5, 6, 9); among these, the cholecystokinin (CCK) system is one of the most extensively characterized (6, 9). In rodents, CCK signaling plays a crucial role in facilitating nociception by counteracting the opioidergic systems in the periaqueductal gray (PAG)—rostral ventral medulla (RVM)—spinal descending pathway (10–13) and inhibiting nociception through the central amygdala (CeA)—PAG—spinal pathway (14). In humans, CCK signaling has been implicated in nocebo hyperalgesia, mediated by the descending nociceptive control system (15). The CCK system is very well-conserved and has been implicated in several common physiological functions among bilaterian species (16–20). However, whether CCK is functionally involved in regulating nociception outside of mammals remains unknown.

Drosulfakinin (DSK), a neuropeptide homologous to CCK, was identified in the fruit fly Drosophila melanogaster (19). Fly DSK is reportedly involved in modulating many physiological functions shared with mammals, including gut functions, anxiety, aggression, memory, feeding, synaptic functions, and courtship behaviors (19, 21–23). After the discovery of stereotyped nociceptive escape behavior called rolling and polymodal Class IV md (C4da) nociceptors (24, 25), the larval Drosophila has been successfully utilized to identify evolutionarily conserved and previously uncharacterized molecular pathways in nociception (26–34) and circuitry mechanisms in the ventral nerve cord (VNC; the invertebrate equivalent of the spinal cord) to compute multimodal sensory stimuli and select nociceptive escape strategies (35–40). Previous studies have demonstrated that neuropeptidergic systems also participate in regulating nociception in Drosophila (39, 41–44). However, the role of fly DSK in nociception remains elusive.

Here, using a collective approach of genetic, behavioral, histological, and Ca2+ imaging analyses, we pursued the mechanisms of DSK-mediated nociceptive regulation and demonstrate that the DSK system constitutes a descending inhibitory pathway of nociception from the brain to the VNC in larval Drosophila.

Results

DSK signaling negatively regulates thermal nociception

Through a thermal nociception screen using the nocifensive rolling response of Drosophila larvae, we found that a deletion mutant line of the dsk gene showed thermal hypersensitivity with a significantly shorter latency in their response to a 42 °C probe than the controls, suggesting that DSK plays a role in negatively regulating nociception (Fig. 1A and B). A genomic fragment containing the wild-type dsk gene, and no other neuropeptide genes, significantly rescued the thermal hypersensitivity of the dsk mutants (Fig. 1B), confirming that dsk is responsible for the thermal hypersensitivity.

DSK signaling is involved in negatively regulating nociception.

(A) Predicted amino acid sequences of pro-DSK peptide in the wild-type (top) and dsksk1 mutants (bottom). Due to a 5-base deletion at the +94-99 position in the coding sequence, dsksk1 mutants are predicted to produce a largely truncated pro-DSK peptide unable to be processed to active DSK peptides. Red letters represent residues different from the wild-type and asterisks indicate stop codons. (B) Hypersensitivity of dsksk1 mutants to a 42°C thermal probe. Significantly shortened latencies of dsksk1 mutants (n = 143) compared with the controls (n = 164) were recovered in genomic rescue animals (Dp(3;1)2-2; dsksk1, n = 139). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. (C) Predicted amino acid sequence of CCKLR-17D1 in the wild-type and CCKLR-17D1 mutants. CCKLR-17D1Δ1, CCKLR-17D1Δ2, and CCKLR-17D1Δ3 have 17-base (+1573-1589), 2-base (+1575-1576) and 32-base (+1573-1604) deletions in the coding sequence respectively, which result in large frameshifts completely abolishing the sixth and seventh transmembrane domains (TMs). (D) CCKLR-17D1Δ1 (n = 54), CCKLR-17D1Δ2 (n = 51) and CCKLR-17D1Δ3 mutants (n = 57) all showed significantly shorter latencies than control (n = 38). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. (E) Predicted amino acid sequence of CCKLR-17D3 in the wild-type and CCKLR-17D3 mutants. CCKLR-17D3Δ1, CCKLR-17D3Δ2, and CCKLR-17D3Δ3 possess 8-base (+474-481), 32-base (+457-488), and 5-base (+472-476) deletions in the coding sequence respectively, which result in large frameshifts and truncation in the middle of the second TM. (F) CCKLR-17D3Δ1 (n = 32), CCKLR-17D3Δ2 (n = 38), and CCKLR-17D3Δ3 mutants (n = 36) all exhibited significantly shorter latencies than control (n = 70). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points.

DSK has been shown to activate two G-protein coupled receptors, CCKLR-17D1 and CCKLR-17D3 (45, 46), which are orthologous to the mammalian CCK receptors, CCKAR (also known as CCK1) and CCKBR (also known as CCK2) (16–18). To test whether these receptors mediate DSK signaling in nociception, we generated deletion mutants for CCKLR-17D1 and CCKLR-17D3 using CRISPR/Cas9 genome editing and tested them for thermal nociception (Fig. 1C-F). When stimulated with a 42 °C probe, three independent deletion lines of the CCKLR-17D1 and CCKLR-17D3 exhibited thermal hypersensitivity (Fig. 1D and F), further supporting the role of DSK signaling in the negative regulation of thermal nociception. To examine whether the phenotypes of CCKLR-17D1 and CCKLR-17D3 mutants are classified as hyperalgesia (hypersensitivity to normally noxious stimuli) or allodynia (abnormal hypersensitivity to normally innocuous stimuli), we tested the receptor mutants with a 38 °C probe, which is close to the threshold of larval thermal nociception (39 °C) (24). We found that the responses of the DSK receptor mutants were indistinguishable from the controls, indicating that the DSK receptor mutants are hypersensitive to suprathreshold thermal stimuli, thus hyperalgesic (Fig. S1).

Two groups of brain neurons expressing DSK are responsible for regulating nociception

We attempted to identify DSK-expressing cells in the larval CNS that are responsible for regulating nociception. Unlike mammalian CCK, which is expressed in the CNS and gastrointestinal system, Drosophila DSK is expressed in the CNS but not in the gut (19, 47–50). Previous immunohistochemical studies have reported putative DSK-expressing cells in the larval CNS (47, 49). However, since the specificity of the DSK antibodies has not been validated with null mutants in the former studies, it has been a concern that the reported DSK-expressing cells could include non-DSK cells that express the other neuropeptides sharing the C-terminal RFa motif with DSK (51). Consistent with this concern, we found that an antibody against crustacean FLRFa, an FMRFa-like neuropeptide with the C-terminal RFa motif, gives rise to a comparable staining pattern to that of the previously reported DSK antibodies, visualizing cells designated as insulin-producing cells (IPCs; referred to as SP3 in Nichols and Lim, (1996) (47)), MP1, SP1, SP2, Sv, SE2, Tv1-3, T2dm, and A8 in the larval CNS (Fig. 2A) (47, 49).

DSK is expressed in two groups of larval brain neurons.

(A) Representative image of DSK-GAL4 expression and anti-FLRFa staining in the wild-type larval CNS. (B) Representative image of anti-FLRFa staining in dsksk1 larval CNS. (C) Representative images of DSK-GAL4 expression and anti-FLRFa staining in the control larval brains. Arrows and arrowheads indicate MP1 and Sv neurons, respectively. Co-expressions of DSK-GAL4 and anti-FLRFa were observed in MP1 and Sv in all of the examined samples (n = 27/27), but in IPCs only in 7% (n = 2/27). (D) Representative images of DSK-GAL4 expression and anti-FLRFa staining in the dsksk1 larval brains. Arrows and arrowheads indicate MP1 and Sv neurons, respectively. (E) Representative images of DSK-GAL4 expression and anti-FLRFa staining in the larval brains of the genomic rescue genotype (Dp(3;1)2-2/Y; +/lexA-rCD2::RFP UAS-mCD8::GFP; dsksk1 DSK-GAL4/dsksk1). Arrows and arrowheads indicate MP1 and Sv neurons, respectively. (F) An example image showing the expression of DSK-2A-GAL4, a 2A-GAL4 knock-in line of the dsk gene, more faithfully recapitulating its endogenous expression. Arrows and arrowheads indicate MP1 and Sv neurons, respectively. MP1 neurons were labeled in 100% of DSK-2AGAL4 samples (n = 41/41), while a single IPC in 51.2% (n = 21/41), multiple IPCs in 24.4% (n = 10/41), and Sv neurons in 17.1% (n = 7/41) of examined samples. All scale bars represent 20 μm.

To identify bona fide DSK-expressing cells responsible for regulating nociception, we performed anti-FLRFa staining in multiple dsk null alleles and found that the staining signals persist in all but two pairs of neurons, MP1 and Sv, in the mutant CNS (Fig. 2B-D and S2A-D), suggesting that these two pairs of neurons are the only neurons in the larval CNS that express DSK. The genomic rescue fragment that rescued the thermal hypersensitivity of dsk mutants (Fig. 1B) also restored the anti-FLRFa signals in MP1 and Sv in a dsk mutant background (Fig. 2E), suggesting that the dsk gene is responsible for the anti-FLRFa signals in these neurons as well as the thermal nociceptive responses of larvae. This expression pattern was further corroborated by the transgenic reporter line DSK-GAL4 and the 2A-GAL4 knock-in reporter line DSK-2A-GAL4 (52), both of which we found were expressed only in MP1 and Sv neurons among the anti-FLRFa-positive neurons (Fig. 2A, C, and F; Fig. S2B-F). We also found that anti-FLRFa, DSK-GAL4, and DSK-2A-GAL4 do not visualize larval peripheral neurons including C4da nociceptors (Fig. 2A and F; Fig. S2G). Taken together, these results identified two sets of brain neurons, MP1 and Sv, as the DSK-expressing cells that are involved in regulating nociception in larvae.

CCKLR-17D1 in Goro neurons functions to negatively regulate nociception

We sought the potential target cells of DSK signaling for regulating nociception. Since DSK receptor mutants were thermally hypersensitive consistently to dsk mutants (Fig. 1B, D, and F), neurons in the larval nociceptive circuit that express DSK receptors were promising candidates. To visualize the cells expressing DSK receptors, we generated T2A-GAL4 knock-ins in CCKLR genes. Both CCKLR-17D1-T2A-GAL4 and CCKLR-17D3-T2A-GAL4 were widely expressed in the larval CNS, predominantly in neuronal cells (Fig. 3A and B; Fig. S3A-D). By performing double-labeling experiments, we found that CCKLR-17D1-T2A-GAL4 and CCKLR-17D3-T2A-GAL4 are not expressed in the nociceptors at the periphery (Fig. 3C and D; Fig. S3E and F), or in nociceptive interneurons including Basin1-4 (53), A08n (36, 41), and DnB neurons (35) in the larval VNC (Fig. S3G-L). However, we found that they are expressed in Goro neurons (Fig. 3E and F), which are the fourth-order nociceptive interneurons located in the larval VNC (53).

DSK receptors are expressed in Goro neurons in the larval VNC.

(A and B) Representative images of CCKLR-17D1-T2A-GAL4 (A) and CCKLR-17D3-T2A-GAL4 (B) expressions in the larval CNS. (C and D) Representative images showing double-labeling of CCKLR-17D1-T2A-GAL4 (C) and CCKLR-17D3-T2A-GAL4 (D) with C4da nociceptors (R38A10-lexA). Arrows and arrowheads indicate cell bodies of C4da nociceptors and es cells, respectively. (E and F) Representative images showing double-labeling of CCKLR-17D1-T2A-GAL4 (E) and CCKLR-17D3-T2A-GAL4 (F) with Goro neurons (arrowheads,

R69E06-lexA). Expression patterns were confirmed in multiple samples (n = 7 and 5). All scale bars represent 20 μm.

To test whether the DSK receptors in Goro neurons are functionally important for regulating nociception, we performed RNAi and rescue experiments using R69E06-GAL4 that marks Goro neurons (53). RNAi knockdown of CCKLR-17D1, but not CCKLR-17D3, in R69E06-GAL4 neurons induced thermal hypersensitivity (Fig. 4A and Fig. S4A). R69E06-GAL4 is also expressed in multiple neurons in the larval brain other than Goro neurons in the VNC (53). However, the thermal hypersensitivity was not observed when the expression of CCKLR-17D1 RNAi was excluded from Goro neurons by tsh-GAL80, pointing to the requirement of CCKLR-17D1 in Goro neurons (Fig. S4B-D). Consistent with these RNAi results, expressing wild-type CCKLR-17D1, but not CCKLR-17D3, with R69E06-GAL4 rescued the hypersensitivity of the respective mutants (Fig. 4B and C). Furthermore, the morphology of Goro neurons was not affected by CCKLR-17D1 knockdown (Fig. S4E). Therefore, we conclude that CCKLR-17D1 functions in Goro neurons to negatively regulate thermal nociception, but the function of CCKLR-17D3 in nociceptive regulations resides elsewhere.

CCKLR-17D1 in Goro neurons is necessary and sufficient for normal thermal nociception.

(A) RNAi of CCKLR-17D1 and CCKLR-17D3 using a Goro-GAL4 line R69E06-GAL4. Expressing CCKLR-17D1 RNAi (R69E06-GAL4 x yv; JF02644, n = 66) but not CCKLR-17D3 RNAi (R69E06-GAL4 x yv; JF02968, n = 95) with R69E06-GAL4 caused significantly shorter latencies to 42 °C than controls (R69E06-GAL4 x yv; attp2, n = 46). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001, NS (non-significant) p > 0.05 Steel’s test. (B) Rescue of CCKLR-17D1 with R69E06-GAL4. The shortened latencies of CCKLR-17D1Δ1 mutants observed in UAS controls (CCKLR-17D1Δ1; UAS-CCKLR-17D1/+, n = 61) or GAL4 controls (CCKLR-17D1Δ1; R69E06-GAL4/+, n = 42) were restored in the rescue genotype (CCKLR-17D1Δ1; UAS-CCKLR-17D1/+; R69E06-GAL4/+, n = 52). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. (C) Rescue of CCKLR-17D3 with R69E06-GAL4. The shortened latencies of CCKLR-17D3Δ1 mutants were unaltered in the rescue genotype (CCKLR-17D3Δ1; UAS-CCKLR-17D3/+; R69E06-GAL4/+, n = 130) compared with UAS controls (CCKLR-17D3Δ1; UAS-CCKLR-17D3/+, n = 40) or GAL4 controls (CCKLR-17D3Δ1; R69E06-GAL4/+, n = 68). (Left) Histograms (Right) Box plots of latencies. NS (non-significant) p > 0.05 Steel’s test. All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points.

Activation of Goro neurons elicits nocifensive rolling in larvae (53). Hence, if CCKLR-17D1 functions in Goro neurons to negatively regulate nociception, the lack of CCKLR-17D1 should sensitize these neurons to noxious heat. We directly addressed this hypothesis by using a Ca2+ imaging technique we had previously developed (32, 35), whereby we monitored GCaMP6m signals in Goro neurons while locally applying thermal ramp stimuli to the larval body wall. In our thermal ramp stimulation protocol, the temperature around the larval CNS hardly reached 28 °C during the heat ramp stimulations (27.0 ± 0.2 °C at the peak, n = 12), suggesting that the larval nociceptive system in the CNS that functions as an internal sensor for fast temperature increase (54) was unlikely to be activated.

Goro neurons in CCKLR-17D1Δ1 mutants showed a significantly steeper GCaMP6m signal increase from 40 to 50 °C in comparison with the wild-type controls (Fig. 5A and B, Movie S1 and S2). Since the baseline fluorescence levels of GCaMP6m were not significantly different in the wild-type and CCKLR-17D1Δ1 mutants (Fig. 5C), these data demonstrate that Goro neurons in CCKLR-17D1Δ1 mutants are specifically sensitized to a noxious range of heat. Suppressing CCKLR-17D1 by RNAi in Goro neurons also induced significantly sensitized responses of Goro to noxious temperatures of 44–49 °C (Fig. 5D-F, Movie S3 and S4). In contrast, Goro neurons in CCKLR-17D3Δ1 mutants exhibited GCaMP6m signals that were mildly elevated but largely parallel compared with that of the controls (Fig. S5, Movie S5 and S6), providing further evidence for the major functioning of CCKLR-17D3 in nociceptive regulation outside Goro neurons. Considered together, these data demonstrate that CCKLR-17D1 functions to negatively regulate the activity of Goro neurons, thereby attenuating behavioral nociceptive responses.

Goro neurons lacking CCKLR-17D1 show sensitized responses to noxious heat.

(A) Representative still images showing thermal activation of Goro neurons in the controls (top, yw/Y; R69E06-GAL4 UAS-GCaMP6m/+) and CCKLR-17D1Δ1 mutants (bottom, CCKLR-17D1Δ1/Y; R69E06-GAL4 UAS-GCaMP6m/+). See also Movie S1 and S2. (B) Average percent increase of GCaMP6m fluorescence intensity relative to the baseline (ΔF/F0) during heat ramp stimulations in controls (n = 15) and CCKLR-17D1Δ1 (n = 15). * p < 0.05, ** p < 0.01 Mann-Whitney’s U-test. (C) Basal GCaMP6m signal levels (F0) in controls (n = 15) and CCKLR-17D1Δ1 mutants (n = 15). p > 0.6 Mann-Whitney’s U-test. (D) Representative stills showing thermal activation of the controls (top, R69E06-GAL4 UAS-GCaMP6m x yv; attp2) and Goro neurons expressing CCKLR-17D1 RNAi (bottom, R69E06-GAL4 UAS-GCaMP6m x yv; JF02644). See also Movie S3 and S4. (E) Average percent increase of GCaMP6m fluorescence intensity relative to the baseline (ΔF/F0) during heat ramp stimulations in controls (n = 15) and CCKLR-17D1 RNAi (n = 16). * p < 0.05 Mann-Whitney’s U-test. (F) Basal GCaMP6m levels (F0) in controls (n = 15) and CCKLR-17D1 RNAi (n = 16). p > 0.6 Mann–Whitney’s U-test. Error bars represent standard error. All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points. All scale bars represent 20 μm.

Neuronal projections and thermal responsiveness of DSK-expressing neurons

If CCKLR-17D1 is involved in regulating the activity of Goro neurons, how can DSK be conveyed from the brain to the VNC? We noticed that some of the DSK-GAL4 positive brain neurons sent descending neural processes to the VNC (Fig. 6A), which were all anti-FLRFa positive (Fig. 6B). Furthermore, the anti-FLRFa signals in the descending projections were completely absent in the dsk mutants (Fig. 6C), suggesting that these descending projections likely originated from the DSK-expressing brain cells, namely MP1 and/or Sv. In analyzing DSK-2A-GAL4, we found that MP1 neurons in fact sent the descending projections to the VNC (Fig. 6D). To further understand the projection patterns of MP1 and Sv neurons in detail, we performed single-cell labeling using an FLP-out technique and revealed that MP1 neurons sent descending projections contralaterally to the VNC (Fig. 6E), while Sv neurons projected within the brain (Fig. 6F). In comparison with the longitudinal processes to the VNC, MP1 neural processes in the brain possessed few anti-FLRFa positive puncta, which represent DSK in these neurons (Fig. 6G). Furthermore, when the somatodendritic marker UAS-Denmark (55) and the synaptic vesicle marker UAS-syt::eGFP were expressed in MP1 neurons, Denmark preferentially localized in the neural processes within the brain while syt::eGFP strongly accumulated in the processes descending to the VNC (Fig. 6H). Collectively, these data demonstrate that MP1 neurons project DSK-positive descending axons to the VNC from the brain.

Projection patterns of DSK-expressing neurons in the larval CNS.

(A) Representative image showing descending axons from DSK-GAL4 positive brain neurons (green) to the larval VNC. (B) Representative image showing all descending projections labeled by DSK-GAL4 (green) harbor punctate anti-FLRFa signals (magenta) in the wild-type. (C) Representative image showing the complete absence of punctate anti-FLRFa signals (magenta) in the descending projections labeled by DSK-GAL4 (green) in the dsksk1 mutants. (D) Image showing the descending projections of MP1 neurons (arrows) marked by DSK-2A-GAL4. (E) Image showing a single FLP-out clone of MP1 neuron (arrow) contralaterally sending descending axons to the larval VNC. The yellow dashed line indicates the midline. (F) Image showing two FLP-out clones of Sv neurons (arrowheads) sending ascending axons to both sides of the brain. (G) A projection image showing the MP1 neurites in the brain. Note that few anti-FLRFa-positve puncta are associated with MP1 neurites within the brain. Similar expression patterns were observed in multiple samples (n = 3). The arrow indicates the anti-FLRFa-positve MP1 soma. (H) Representative image showing localizations of syt::eGFP (green) and Denmark (magenta) in MP1 neurons (DSK-2A-GAL4 x UAS-syt::eGFP UAS-Denmark). Similar expression patterns were observed in multiple samples (n = 7). All scale bars represent 20 μm.

Next, we investigated whether some of these DSK-expressing neurons are responsive to noxious thermal stimuli. In order to examine the responsiveness of multiple neurons in intact larvae, we took the snapshot approach using CaMPARI, a genetically-encoded Ca2+ sensor whose fluorescence irreversibly changes from the green to the red region of the spectrum in response to irradiation with 405 nm UV light in a Ca2+ dependent manner, and is thus useful to monitor the activities of multiple neurons simultaneously in intact animals (56). When the larvae expressing CaMPARI2 (an improved version of CaMPARI (57)) in Goro neurons were stimulated by local thermal ramp stimuli under a 405 nm UV light, a significantly increased CaMPARI2 photoconversion was detected compared to the control group that had only been irradiated with 405nm UV (Fig. 7A and B), confirming that by using CaMPARI2 in our experimental setup we managed to successfully capture the increased activity of Goro neurons in response to noxious heat in intact larvae. We then tested the responsiveness of DSK-2A-GAL4 neurons using the same stimulation and photoconversion protocol and found that none of the MP1 or Sv neurons, nor IPCs showed increased CaMPARI2 photoconversion in response to local thermal ramp stimuli (Fig. 7A, C-E). Interestingly, MP1 neurons exhibited high CaMPARI2 photoconversion regardless of the presence of local heat ramp stimuli, which was even comparable to that observed in Goro neurons activated by noxious heat stimuli (Fig. 7A and C). These data suggest that none of the DSK-2A-GAL4 neurons are clearly responsive to noxious heat, but the markedly high CaMPARI2 photoconversion observed in MP1 neurons raises the possibility that these neurons may be active regardless of the presence of noxious heat stimuli.

Activity and thermal responsiveness of DSK-expressing neurons

(A) Box plots of the quantified ratio of CaMPARI2 green signals and photoconverted CaMPARI2 red signals detected by anti-CaMPARI-red in Goro, MP1, Sv, and IPCs (Red/Green) with or without local heat ramp stimulation. Goro neurons showed a significantly higher Red/Green ratio of CaMPARI2 when local heat ramp stimulation was applied to the larvae (n = 16 and 20, p < 0.001, Mann-Whitney’s U-test). MP1 neurons showed a high Red/Green ratio of CaMPARI2 both with and without local heat stimulation (n = 21 and 20, p > 0.3, Mann-Whitney’s U-test). The Red/Green ratio of CaMPARI2 was low in Sv compared to MP1 neurons, and CaMPARI2 photoconversion was not increased by local heat stimulation (n = 9 and 10, p > 0.5, Mann-Whitney’s U-test). IPCs also exhibited low CaMPARI2 photoconversion regardless of the presence of local heat ramp stimulation (n = 17 and 16, p > 0.25, Mann-Whitney’s U-test). (B-E) Representative images showing CaMPARI2 green signals (green) and anti-CaMPARI-red signals (magenta) in Goro (B), MP1 (C), Sv (D), and IPCs (E). All box plots show median (middle line) and 25th to 75th percentiles, with whiskers indicating the smallest to the largest data points. Scale bars represent 10 μm.

Connectivity between DSKergic and Goro neurons

To gain more insights into the connectivity between DSK-expressing brain neurons and Goro neurons, we further examined the anatomical relationship between MP1 axons and Goro neurons in the VNC. Performing double-labeling experiments, we found that MP1 axons with anti-FLRFa signals (representing DSK in MP1 axons) partially overlapped with Goro neurites in the thoracic segments of the larval VNC (Fig. 8A). GFP reconstitution was reproducibly detected between MP1 and Goro neurons using the CD4-GRASP (GFP Reconstitution Across Synaptic Partners) system (58, 59), thus confirming that MP1 axons and Goro neurons are indeed in close proximity (Fig. 8B). Since the CD4-GRASP is known to detect general cell-cell contacts (60, 61), we further investigated whether MP1 and Goro neurons are synaptically connected by using the nSyb-GRASP technique, which specifically enables synapse detection by localizing one of the split-GFP fragments in presynaptic sites (62). However, no signals of reconstituted GFP were detected between MP1 and Goro with nSyb-GRASP, although we found proximate localizations of nSyb-GFP1-10 in the MP1 axons with the Goro neurites expressing CD4::GFP11 (Fig. 8C). We also employed the trans-Tango system (63) as another tool to detect synaptic connectivity. However, when the trans-Tango ligand was expressed by DSK-2A-GAL4, synaptic connectivity between MP1 and Goro was not indicated since the postsynaptic activation of trans-Tango signals was not observed in Goro neurons, although some trans-Tango-positive neurons were observed around Goro neurons (Fig. 8D).

Synaptic connectivity between MP1 axons and Goro neurons is not detected

(A) Representative images of the larval VNC showing partially overlapping MP1 axons (green) and processes of Goro neurons (blue) that are associated with anti-FLRFa signals (red). MP1 axons partially overlapping Goro neurons were similarly observed in all examined samples (n = 5). (B) Representative image showing CD4-GRASP experiments between MP1 axons and Goro neurons (magenta). Signals of reconstituted GFP (GRASP, green) were detected in all examined samples (n = 5). (C) Representative image showing nSyb-GRASP experiments between MP1 axons (blue) and Goro neurons (red). Signals of reconstituted GFP (GRASP, green) were not detectable in any of the examined samples (n = 5). (D) Representative image showing trans-Tango experiments using DSK-2A-GAL4 (R69E06-lexA, lexAop-rCD2::RFP UAS-mCD8::GFP; DSK-2A-GAL4 x UAS-myrGFP, QUAS-mtdTomato::3xHA; trans-Tango). Transsynaptic activation of Tango signals (blue; mtdTomato::3xHA detected by anti-HA) was undetectable in Goro neurons (red; rCD2::RFP detected by anti-CD2). Green represents MP1 axons labeled by DSK-2A-GAL4 (myrGFP and mCD8::GFP detected by anti-GFP). Signals of mtdTom::3xHA in Goro neurons were not detected in any of the examined samples (n = 4). Arrows indicate Goro somata. All scale bars represent 20 μm, except for the insets in (A) (5 μm).

Although we could not find positive evidence for synaptic connectivity between MP1 axons and Goro neurons, we further tested the hypothesis that DSK-expressing neurons in the brain may have functional connectivity to Goro neurons. To achieve this, we first attempted artificial activation experiments of DSK-2A-GAL4 neurons using UAS-dTRPA1 to determine whether activating DSK-expressing neurons actually attenuates larval nociception. dTRPA1 is a cation channel gated by warm temperature (> 29 °C) that has been used as a tool to artificially activate neurons of interest in Drosophila (64). When the larvae expressing dTRPA1 in C4da nociceptors with ppk1.9-GAL4 were placed in a chamber at 35°C, dTRPA1-induced thermogenetic activation of C4da nociceptors caused nocifensive rolling responses within two seconds in the majority of animals, as reported previously (65) (Fig. 9A and B). In contrast, when the larvae expressing dTRPA1 simultaneously in ppk1.9-GAL4 and DSK-2A-GAL4 were placed in a chamber at 35°C, the additional activation of DSK-2A-GAL4 neurons to C4da nociceptors resulted in a markedly reduced percentage of larvae showing rolling responses within two seconds and in significantly lengthened latencies to exhibit a rolling response (Fig. 9A and B). None of the animals expressing dTRPA1 with DSK-2A-GAL4 alone showed a rolling response within 10 seconds (Fig. 9A).

Thermogenetic activation of DSK-2A-GAL4 neurons inhibits larval nociception and Goro neurons

(A and B) Thermogenetic activations of nociceptors and/or DSK-2A-GAL4 neurons using UAS-dTRPA1. (A) Histogram representing the percentage of larvae exhibiting nociceptive rolling. Upon thermogenetic activations, 55% of larvae expressing dTRPA1 in nociceptors (ppk-GAL4 > dTRPA1; ppk1.9-GAL4 x UAS-dTRPA1, n = 58) exhibited nociceptive rolling within 2 seconds. In contrast, only 11% of larvae with simultaneous activations of nociceptors and DSK-2A-GAL4 showed rolling within 2 seconds (ppk-GAL4 & DSK-2A-GAL4 > dTRPA1; ppk1.9-GAL4 DSK-2A-GAL4 x UAS-dTRPA1, n = 54). Thermogenetic activations of DSK-2A-GAL4 neurons alone (DSK-2A-GAL4 > dTRPA1; DSK-2A-GAL4 x UAS-dTRPA1, n = 40) did not trigger nociceptive responses even after 10 seconds. (B) Box plots showing latencies. Thermogenetic activation of DSK-2A-GAL4 simultaneously with nociceptors caused a significantly longer latency to rolling responses (Mann-Whitney U-test, *** p < 0.001). (C and D) CaMPARI2 snapshot imaging of the activity of Goro neurons activated by thermogenetic stimulation of C4da nociceptors. (C) Box plots of the ratio of CaMPARI2 green signals and photoconverted CaMPARI2 red signals detected by anti-CaMPARI-red in Goro neurons (Red/Green). Compared to the larvae with thermogenetic activation of C4da nociceptors only (ppk-GAL4 > dTRPA1; ppk1.9-GAL4, UAS-dTRPA1 x R69E06-lexA; lexAop-CaMPARI2; n = 17), the larvae with simultaneous activation of C4da and DSK-2A-GAL4 neurons (ppk-GAL4 & DSK-2A-GAL4 > dTRPA1; ppk1.9-GAL4, UAS-dTRPA1; DSK-2A-GAL4 x R69E06-lexA; lexAop-CaMPARI2; n = 19) showed significantly decreased CaMPARI2 photoconversion in Goro neurons (Mann-Whitney U-test, * p < 0.025). (D) Representative images showing CaMPARI2 green signals (green) and anti-CaMPARI-red signals (magenta) in Goro neurons of animals with thermogenetic C4da activation (ppk-GAL4 > dTRPA1) and with simultaneous thermogenetic activation of C4da and DSK-2A-GAL4 neurons (ppk-GAL4 & DSK-2A-GAL4 > dTRPA1). All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points. Scale bars represent 10 μm.

The above data suggest that activation of DSK-2A-GAL4 neurons inhibits larval nociception. We then investigated whether activating DSK-2A-GAL4 neurons could attenuate the activity of Goro neurons, by combining the artificial activation experiments described above with CaMPARI2 snapshot imaging of Goro neurons. Indeed, CaMPARI2 photoconversion in Goro neurons induced by the thermogenetic activations of C4da nociceptors was significantly decreased by 35% with the simultaneous activation of DSK-2A-GAL4 neurons (Fig. 9C and D). Given that MP1 neurons were the only cells that were labeled by DSK-2A-GAL4 with 100% reproducibility, while the expression of DSK-2A-GAL4 in IPCs and Sv neurons was minor and stochastic (Fig. 2F), the observed inhibition of behavioral nociception and Goro neurons caused by activating DSK-2A-GAL4 neurons is mostly attributable to the activation of MP1 neurons. Thus, these data support the functional connectivity between DSKergic MP1 neurons in the brain and Goro neurons.

Discussion

In this study, we have demonstrated that (1) DSK and its receptors CCKLR-17D1 and CCKLR-17D3 are involved in negatively regulating thermal nociception, (2) Two sets of brain neurons, MP1 and Sv, express DSK in the larval nervous system, (3) One of the DSK receptors (CCKLR-17D1) functions in Goro neurons in the VNC to negatively regulate thermal nociception, and (4) Thermogenetic activation of DSK-2A-GAL4 neurons attenuates both larval nociception and the activity of Goro neurons. Based on these results, we propose that the DSKergic neurons regulating the activity of Goro neurons constitute a descending inhibitory pathway of nociception from the brain to the VNC in larval Drosophila (Fig. 10). To our knowledge, our findings represent the first evidence of a descending mechanism modulating nociception from the brain in a non-mammalian species.

A schematic model of the DSKergic descending inhibitory pathway of nociception in larval Drosophila.

DSK-expressing MP1 and/or Sv neurons in the brain secrete DSK peptides. DSK in the larval VNC activates the CCKLR-17D1 receptor expressed in Goro neurons, which subsequently inhibits the activity of Goro neurons, and ultimately larval nociceptive rolling responses.

DSKergic descending system as a physiological modulator of nociception

DSK has been implicated in multiple physiological and developmental processes in Drosophila (19, 21–23). Previous studies have shown that dsk and CCKLR-17D1 mutants exhibit significant reductions of synaptic growth and excitability in larval neuromuscular junctions (NMJ) and larval locomotion under bright light (45, 66), suggesting the importance of the DSK/CCKLR-17D1 signaling pathway in the developmental processes of motoneurons. However, in this study, no major developmental defects were observed in Goro neurons with CCKLR-17D1 RNAi (Fig. S4E). Furthermore, the simultaneous thermogenetic activations of DSK-2A-GAL4 neurons and C4da nociceptors inhibited larval nociception and the activity of Goro neurons (Fig. 9). Given the pronociceptive role of Goro neurons, their reduced synaptic or neuronal activity should cause nociceptive insensitivity or reduced Ca2+ responses, which is contradicting to our observation that CCKLR-17D1 knockdown in Goro produced the thermal hypersensitivity and exaggerated Ca2+ responses (Fig 4 and 5). Thus, these data consistently support a physiological role for DSKergic descending signals in the modulation of the activity of Goro neurons rather than a developmental role, highlighting the functional differences of the DSK/CCKLR-17D1 pathway between the NMJ and the nociceptive system in the CNS.

DSK in larval IPCs dispensable for nociceptive modulation

A previous study has reported the expression of DSK in a population of larval IPCs and its functions in responding to starving conditions (49). However, we observed that anti-FLRFa signals in IPCs persisted in dsk null mutants and that anti-FLRFa signals in IPCs hardly overlap with either DSK-GAL4 or DSK-2A-GAL4 expressions (Fig. 2 and S2). Although we used a different antibody from that in the study mentioned above, the number of IPCs visualized by anti-FLRFa was comparable to that of cells visualized by anti-DSK (Fig. S2) (49). Regarding the functions of IPCs in nociception, Im et al. showed that silencing dilp2-GAL4 positive IPCs has no significant effect on the baseline nociceptive responses (67). In our CaMPARI2 imaging, IPCs marked by DSK-2A-GAL4 showed very low baseline activity and no responsiveness to noxious heat (Fig. 7). Thus, the current and previous studies strongly suggest that IPCs are likely irrelevant for DSK-mediated nociceptive regulation, at least under normal conditions, although it is still possible for DSK to be expressed stochastically in a limited population of IPCs.

Functional mechanisms of DSKergic nociceptive inhibitory system

Although our data suggest the existence of a DSKergic descending inhibitory system from the brain to the Goro neurons in the VNC, there are multiple remaining questions for future studies on the mechanisms by which DSK-expressing neurons in the brain regulate Goro neurons: First, whether both MP1 and Sv neurons are involved in nociceptive inhibition remains unclear. Our findings consistently point to MP1 neurons over Sv neurons as a major source of DSK regulating Goro neurons. However, given the potential of DSK for distant actions through diffusion, the possibility of Sv neurons communicating with Goro neurons cannot be eliminated. Further analysis using finer genetic/neuronal tools will be necessary to tease the functions of MP1 and Sv neurons apart and understand the functional mechanism of the DSKergic system in the regulation of larval nociception.

Second, how this DSKergic system is activated to regulate nociception in vivo needs further investigation. We observed very high CaMPARI2 photoconversion in MP1 neurons without noxious heat stimuli, which did not further increase in response to noxious heat application (Fig. 7). These data raise the hypothesis that MP1 neurons may function as a tonically active brake for nociceptive circuits, rather than a simple negative feedback system that is activated by nociceptive stimuli as input. The tonically active model of MP1 function is also consistent with our data showing that dsk and CCLR-17D1 mutations, as well as CCLR-17D1 RNAi in Goro neurons, all caused thermal hypersensitivity under normal conditions (Fig. 1 and 4). Up to this point, no external or internal signals have been directly shown to be the input into larval DSKergic neurons. However, since DSK has been implicated in the regulation of several physiological processes such as hunger and stress, sensory or molecular signals that are involved in these physiological processes have the potential to serve as upstream cues that activate DSKergic neurons. Alternatively, it is also possible that MP1 neurons may have spontaneous activity. Further physiological studies on the activity and responsiveness of DSKergic neurons would be required to elucidate how the DSKergic system functions to regulate nociception in larvae under normal conditions.

Third, the transmission mechanisms of DSK from the brain to Goro neurons need further clarification. Our nSyb-GRASP and trans-Tango experiments consistently showed negative results, indicating no synaptic connectivity between MP1 descending axons and Goro neurons (Fig. 8). Thus, the interaction between DSK from DSKergic neurons in the brain and Goro neurons in the VNC may be mediated non-synaptically through volume transmission, as described in many neuropeptidergic systems (68). Many of the CCKLR-expressing neurons in the VNC are distantly located from the descending MP1 axons (Fig. 3 and 6). Since it is unlikely that all these CCKLR-expressing neurons are synaptically connected to MP1 axons, neuronal communications through volume transmission can be fairly assumed for DSKergic systems in the larval VNC. More detailed analyses at an electron microscopic level of the circuitry connectivity between MP1 and Goro neurons and of the localization of DSK-containing vesicles in DSKergic neurons as well as DSK receptors in Goro neurons would be necessary to further clarify the transmission mechanisms of DSK from the brain to Goro neurons.

Multiple molecular pathways for DSK signaling in the regulation of nociception

The data presented in this study also suggest that DSK signaling could regulate larval nociception through multiple pathways other than the DSK/CCKLR-17D1 system. For example, while CCKLR-17D3 mutants exhibited as severe thermal hypersensitivity as CCKLR-17D1 mutants (Fig. 1D and F), our RNAi and rescue experiments failed to locate the function of CCKLR-17D3 to Goro neurons (Fig. 4A and C). Our GCaMP Ca2+ imaging also revealed that Goro neurons lacking CCKLR-17D3 showed modestly sensitized responses to noxious heat (Fig. S5). These results suggest the major functioning of CCKLR-17D3 in Goro-independent nociceptive pathways. We also observed that DSK receptor mutants exhibited more severe thermal phenotypes than dsk mutants (Fig. 1), which may indicate the presence of unidentified DSK- or CCKLR-dependent signaling pathways in nociception-related cells. Further research is apparently necessary to reveal the whole picture of nociceptive regulations mediated by DSK and its receptors in larval Drosophila.

Potential conservation of CCK-mediated descending nociceptive controls

The CCK system is thought to be one of the most ancient neuropeptide systems, suggested to have multiple common physiological functions across taxa (16–20). In this study, we demonstrate that CCKergic signaling in Drosophila participates in nociceptive modulation through a descending inhibitory pathway similarly to the mammalian CCK system, adding new evidence about the conserved physiological roles of CCK.

Unlike the peripheral nociceptive systems, it is still challenging to align the Drosophila and mammalian CCKergic descending pathways due to low homologies of the CNS structures between Drosophila and mammals, wide-spread CCK expression in the mammalian CNS, and the multiple roles of the CCKergic systems in mammalian nociceptive controls (10–14, 69–73). However, the common usage of an orthologous molecular pathway in descending controls of nociception between the two evolutionarily distant clades raises a fascinating new hypothesis that the descending control from the brain may also be an ancient, conserved mechanism of nociception, which has emerged in the common ancestor of protostomes and deuterostomes. It will be of interest for future research to investigate whether the role of CCK signaling in descending nociceptive controls is also present in other species.

Potentials of non-mammalian models to study the mechanisms of pain modulation and pain pathology in the CNS

Non-mammalian model systems have been increasingly recognized as powerful tools to identify novel pain-related molecular pathways (2, 74–77). However, the utilization of these models has so far been mostly limited to the research on peripheral pain pathophysiology, and few studies have used them to investigate central pain pathophysiology (2, 78). Descending nociceptive control mechanisms are crucial for central pain modulation and have been implicated in the development of chronic pain states in humans (5, 6). Thus, the current study opens the door to a new approach to using powerful neurogenetic tools and the simpler nervous system of Drosophila for elucidating the functional principles of descending nociceptive systems, which may potentially contribute to our understanding of the mechanisms underlying central pain modulation and pain pathology due to dysfunctions of descending modulatory pathways.

Supporting information

eLife Assessment

This is a very interesting and important study that convincingly demonstrates a descending pathway for the control of nociception in non-mammalian organisms.

Reviewer #1 (Public Review):

The study by Oikawa and colleagues demonstrates for the first time that a descending inhibitory pathway for nociception exists in non-mammalian organisms, such as Drosophila. This descending inhibitory pathway is mediated by a Drosophila neuropeptide called Drosulfakinin (DSK), which is homologous to mammalian cholecystokinin (CCK). The study creates and uses several Drosophila mutants to convincingly show that DSK negatively regulates nociception. They then use several sophisticated transgenic manipulations to demonstrate that a descending inhibitory pathway for nociception exists in Drosophila.

Strengths:

This study creates the possibility of using Drosophila to study descending nociceptive systems.

CRISPR/Cas9 is used to generate mutants of dsk, CCKLR-17D1, and CCKLR-17D3. The authors then use these mutants to clearly show that DSK negatively regulates nociception.

Several GAL4s are used to clearly show that these effects are likely mediated by two sets of neurons in the brain, MP1 and Sv.

RNAi and rescue experiments further show that CCKLR-17D1, a DSK receptor, functions in Goro neurons to negatively regulate nociception.

Thermogenetic experiments nicely show that activation of DSK neurons attenuates the nociceptive response.

Weaknesses:

A minor weakness in the study is that it is unclear how DSK negatively regulates nociception. An earlier study at the Drosophila nmj shows that loss of DSK signaling impairs neurotransmission and synaptic growth. In the current study, loss of CCKLR-17D1 in Goro neurons seems to increase intracellular calcium levels in the presence of noxious heat. An interesting future study would be the examination of the underlying mechanisms for this increase in intracellular calcium.

Reviewer #2 (Public Review):

This is an exceptional study that provides conclusive evidence for the existence of a descending pathway from the brain that inhibits nociceptive behavioral outputs in larvae of Drosophila melanogaster. The authors identify molecular both molecular and neuronal/cellular components of this pathway. Converging lines of evidence and conclusive genetic experiments indicate that the neuropeptide, drosulfakinin (DSK), and its receptors (CCK1 and CCK2) function to inhibit nociception behaviors. Interestingly, the authors show that the relevant DSK neurons have cell bodies that are in the larval brain and that these neurons send projections into the thoracic ganglion and ventral nerve cord. Several lines of evidence support the hypothesis that fourth-order nociceptive neurons called Goro, are one relevant target for these outputs. RNAi knockdown of the CCK1 receptor in these cells sensitizes behavioral and physiological responses to noxious heat. Second, the axons of DSK neurons form physical contact with processes of Goro neurons as revealed by GRASP analysis. However, the authors' careful experiments indicate that the contacts between axons and Goro neurites might not be indicative of direct synapses and instead might operate through the bulk transmission of the peptidergic signals. The study raises many interesting questions for future study such as what behavioral contexts might depend on this pathway. Using the CAMPARI approach, the authors do not find that the DSK neurons are activated in response to nociceptive input but instead suggest that these cells may be tonically active in gating nociception. Future studies may find contexts in which the output of the DSK neurons is inhibited to facilitate nociception, or contexts in which the cells are more active to inhibit nociception.

Reviewer #3 (Public Review):

This study describes a descending circuit that can modulate pain perception in the drosophila larvae. While descending inhibition is a major component of mammalian pain perception, it is not known if a similar circuit design exists in fruit flies. Overall the authors use clean logic to establish a role for DSK and its receptor in regulating nociception. I have made a few suggestions that I believe would strengthen the manuscript as this is an important discovery.

Major comments:

1. It's not completely clear why the authors are staining animals with an FLRFa antibody. Can the authors stain WT and DSK KO animals with a DSK antibody? Also, can the authors show in supplemental what antigen the FLRFa antibody was raised against, and what part of that peptide sequence is retained in the DSK sequence? This overall seems like a weakness in the study that could be improved on in some way by using DSK-specific tools.

2. What is the phenotype of DSK-Gal4 x UAS-TET animals? They should be hyper-reactive. If it's lethal maybe try an inducible approach.

3. Figure 9. This was not totally clear, but I think the authors were evaluating spontaneous (i.e. TRPA1-driven) rolling at 35C. The critical question is "does activating DSK-expressing neurons suppress acute heat nociception" and this hasn't really been addressed. The inclusion of PPK Gal4 + DSK Gal4 in the same animal kind of clouds the overall conclusions the reader can draw. The essential experiment is to express UAS-dTRPA1 in DSK-Gal4 or GORO-Gal4 cells, heat the animals to ~29C, and then test latency to a thermal heat probe (over a range of sub and noxious temperatures). Basically prove the model in Figure 10 showing ectopic activation or inhibition for each major step, then test heat probe responses.

4. It would also then be interesting to see how strong the descending inhibition circuit is in the context of UV burn. If this is a real descending circuit, it should presumably be able to override sensitization after injury.

Author Response:

The following is the authors' response to the original reviews.

We sincerely thank all the editors and reviewers for taking the time to evaluate this study. Here is our point-by-point response to the reviewers’ comments and concerns.

Reviewer #1 (Public Review):The study by Oikawa and colleagues demonstrates for the first time that a descending inhibitory pathway for nociception exists in non-mammalian organisms, such as Drosophila. This descending inhibitory pathway is mediated by a Drosophila neuropeptide called Drosulfakinin (DSK), which is homologous to mammalian cholecystokinin (CCK). The study creates and uses several Drosophila mutants to convincingly show that DSK negatively regulates nociception. They then use several sophisticated transgenic manipulations to demonstrate that a descending inhibitory pathway for nociception exists in Drosophila.

[…]

Weaknesses:

A minor weakness in the study is that it is unclear how DSK negatively regulates nociception. An earlier study at the Drosophila nmj shows that loss of DSK signaling impairs neurotransmission and synaptic growth. In the current study, loss of CCKLR-17D1 in Goro neurons seems to increase intracellular calcium levels in the presence of noxious heat. An interesting future study would be the examination of the underlying mechanisms for this increase in intracellular calcium.

We thank the reviewer for the kind and very positive evaluation of our manuscript. We agree that this study has not elucidated the intracellular molecular pathway(s) downstream of CCKLR-17D1 that are involved in the regulation of the activity of Goro neurons, and we think that it would definitely be an interesting topic for future research.

Reviewer #1 (Recommendations For The Authors):

The response latencies for the control yw larvae seem large, with many larvae appearing to be insensitive to the thermal stimulus. Is this just an effect of the yw genetic background? A brief discussion of this might be helpful.

We thank the reviewer for pointing this out. We have also noticed that the yw control larvae tend to show longer response latencies than the other control strains, and in the revised manuscript, we have added the following sentence in the Result section (Lines 91–94):

“We have noticed that the yw control strain, which was used by us to generate the dsk and receptor deletion mutants, showed relatively longer response latencies to the 42 °C probe compared to the other control strains used in this study. This may be attributed to the effect of the genetic background, although, presently, the cause for this difference is unknown.”

Reviewer #2 (Public Review):_

This is an exceptional study that provides conclusive evidence for the existence of a descending pathway from the brain that inhibits nociceptive behavioral outputs in larvae of Drosophila melanogaster. […] The study raises many interesting questions for future study such as what behavioral contexts might depend on this pathway. Using the CAMPARI approach, the authors do not find that the DSK neurons are activated in response to nociceptive input but instead suggest that these cells may be tonically active in gating nociception. Future studies may find contexts in which the output of the DSK neurons is inhibited to facilitate nociception, or contexts in which the cells are more active to inhibit nociception._

Reviewer #2 (Recommendations For The Authors):I have no recommendations for the authors as this is a very complete and thoroughly executed study. The writing is crystal clear.

We thank the reviewer for the kind and very positive evaluation of our manuscript. We are happy to know that our current manuscript was deemed to be clear and convincing by the reviewer.

Reviewer #3 (Public Review):[…] Overall the authors use clean logic to establish a role for DSK and its receptor in regulating nociception. I have made a few suggestions that I believe would strengthen the manuscript as this is an important discovery.

Major comments:

1. It's not completely clear why the authors are staining animals with an FLRFa antibody. Can the authors stain WT and DSK KO animals with a DSK antibody? Also, can the authors show in supplemental what antigen the FLRFa antibody was raised against, and what part of that peptide sequence is retained in the DSK sequence? This overall seems like a weakness in the study that could be improved on in some way by using DSK-specific tools.

We thank the reviewer for this query. We would like to clarify that we first tried the FLRFa antibody to visualize an RFamide-type neuropeptide other than DSK in Drosophila and found that the staining pattern is quite similar to that of anti-DSK, as shown by Nichols et al. [1]. According to the original paper describing the anti-FLRFa antisera [2] (already cited in the reviewed manuscript), the antigen used to raise it was the Phe-Met-Arg-Phe-NH2 peptide conjugated with succinylated thyroglobulin, and the study experimentally shows that the antibody well binds to peptides containing Met-Arg-Phe-NH2 or Leu-Arg-Phe-NH2 sequence and has 100% cross-reactivity to FLRFa. As DSK contains Met-Arg-Phe-NH2 sequence [3], the cross-reaction of this antibody to DSK is consistent with the description of the original study.

Although we were unable to use an antibody specific to DSK, our staining data with dsk deletion mutants and the expression pattern of DSK-2A-GAL4 corroborate each other (Figure 2 and Figure 2-figure supplement 1), which we believe provides compelling evidence for the specific expression of DSK in MP1 and Sv neurons, and for that DSK-2A-GAL4 is a reasonably effective tool to specifically manipulate DSK-expressing neurons.

2. What is the phenotype of DSK-Gal4 x UAS-TET animals? They should be hyper-reactive. If it's lethal maybe try an inducible approach.

We thank the reviewer for this question. Unfortunately, we have not attempted this experiment, although we agree that this would be a nice addition to further strengthen the study if TET worked well in the DSKergic neurons.

3. Figure 9. This was not totally clear, but I think the authors were evaluating spontaneous (i.e. TRPA1-driven) rolling at 35C. The critical question is "does activating DSK-expressing neurons suppress acute heat nociception" and this hasn't really been addressed. The inclusion of PPK Gal4 + DSK Gal4 in the same animal kind of clouds the overall conclusions the reader can draw. The essential experiment is to express UAS-dTRPA1 in DSK-Gal4 or GORO-Gal4 cells, heat the animals to ~29C, and then test latency to a thermal heat probe (over a range of sub and noxious temperatures). Basically prove the model in Figure 10 showing ectopic activation or inhibition for each major step, then test heat probe responses.

We thank the reviewer for suggesting ideas for alternative experiments to potentially strengthen our conclusion. Regarding experiments using heat probes, previous studies have demonstrated that (i) Blocking ppk1.9-GAL4-positive C4da neurons almost completely abolishes the larval nociceptive response to local heat stimulations [4]; (ii) Local heat stimuli above 39 °C readily activate C4da neurons and larval nociceptive rolling [5-9]; and (iii) Thermogenetically or optogenetically activating these neurons is sufficient to trigger Goro neurons and larval rolling [4, 10-12]. Thus, it has now been made clear that heat probes induce larval nociceptive rolling via excitation of the C4da pathway, and we believe that our experiments using thermogenetic activation of C4da neurons can be safely interpreted as an alternative to experiments using heat probes. Using heat probes demands a more complicated experimental set-up to be combined with CaMPARI imaging experiments, and this is another reason why we preferred to take the thermogenetic approach.

We have also considered the experiment using Goro-GAL4 instead of ppk-GAL4. However, if dTRPA1 artificially activates Goro neurons far downstream of the neuronal mechanism by which MP1 activation suppresses Goro neuron activity, the effect of MP1 activation may be bypassed and masked. As we currently do not know the epistasis between dTRPA function and the effect of MP1 activation in modulating the activity of Goro neurons, we rather chose to activate C4da neurons by using ppk-GAL4, which likely resulted in more natural activation of Goro neurons than dTRPA1-triggered direct activations.

4. It would also then be interesting to see how strong the descending inhibition circuit is in the context of UV burn. If this is a real descending circuit, it should presumably be able to override sensitization after injury.

Indeed, this is an interesting avenue to explore in future studies to understand the type of situation in which the DSKergic descending system functions to control nociception.

Reviewer #3 (Recommendations For The Authors):Overall this is a good story and the claims are generally supported with experimental evidence. The way to really improve this study would be to use more precise and definitive tools, like specific antibodies, specifically targeted genes, and better temporal control of the descending circuit to prove this is inducible sufficient to suppress acute thermal nociception and this occurs only via a descending pathway, etc. However this would be exponentially more work, and so the authors I guess need to weigh the cost-benefit of definitive proof vs. strong evidence for their claims. Overall I think this study will be the beginning of a new line of inquiry in the field that has the potential to guide our understanding also of mammalian descending pathways, and as such, this study is of value to the community.

We appreciate the reviewer’s multiple interesting ideas for experiments that could have been performed to further reinforce our findings. We agree that some experiments that the reviewer suggested would potentially strengthen this work if supplemented. However, as aforementioned, in our humble opinion, we think that the experiments that the reviewer suggested are either outside the scope of this paper or have no significant benefits over the experiments that were already conducted, and hence are not essential to the present study.

References

1. Nichols, R. and I.A. Lim, Spatial and temporal immunocytochemical analysis of drosulfakinin (Dsk) gene products in the Drosophila melanogaster central nervous system. Cell Tissue Res, 1996. 283(1): p. 107-16.

2. Marder, E., et al., Distribution and partial characterization of FMRFamide-like peptides in the stomatogastric nervous systems of the rock crab, Cancer borealis, and the spiny lobster, Panulirus interruptus. J Comp Neurol, 1987. 259(1): p. 150-63.

3. Nassel, D.R. and M.J. Williams, Cholecystokinin-like peptide (DSK) in Drosophila, not only for satiety signaling. Front Endocrinol, 2014. 5.

4. Hwang, R.Y., et al., Nociceptive neurons protect Drosophila larvae from parasitoid wasps. Curr Biol, 2007. 17(24): p. 2105-2116.

5. Tracey, W.D., Jr., et al., painless, a Drosophila gene essential for nociception. Cell, 2003. 113(2): p. 261-73.

6. Xiang, Y., et al., Light-avoidance-mediating photoreceptors tile the Drosophila larval body wall. Nature, 2010. 468(7326): p. 921-6.

7. Burgos, A., et al., Nociceptive interneurons control modular motor pathways to promote escape behavior in Drosophila. eLife, 2018. 7.

8. Honjo, K. and W.D. Tracey, Jr., BMP signaling downstream of the Highwire E3 ligase sensitizes nociceptors. PLoS Genet, 2018. 14(7): p. e1007464.

9. Im, S.H., et al., Tachykinin acts upstream of autocrine Hedgehog signaling during nociceptive sensitization in Drosophila. eLife, 2015. 4: p. e10735.

10. Ohyama, T., et al., A multilevel multimodal circuit enhances action selection in Drosophila. Nature, 2015. 520(7549): p. 633-9.

11. Honjo, K., R.Y. Hwang, and W.D. Tracey, Jr., Optogenetic manipulation of neural circuits and behavior in Drosophila larvae. Nat Protoc, 2012. 7(8): p. 1470-8.

12. Zhong, L., et al., Thermosensory and non-thermosensory isoforms of Drosophila melanogaster TRPA1 reveal heat sensor domains of a thermoTRP channel. Cell Rep, 2012. 1(1): p. 43-55.

Version 2

Abstract

Abstract

Nociception is a neural process that animals have developed to avoid potentially tissue-damaging stimuli. While nociception is triggered in the peripheral nervous system, its modulation by the central nervous system is a critical process in mammals, whose dysfunction has been extensively implicated in chronic pain pathogenesis. The peripheral mechanisms of nociception are largely conserved across the animal kingdom. However, it is unclear whether the brain-mediated modulation is also conserved in non-mammalian species. Here, we show that Drosophila has a descending inhibitory mechanism of nociception from the brain, mediated by the neuropeptide Drosulfakinin (DSK), a homolog of cholecystokinin (CCK) that plays an important role in the descending control of nociception in mammals. We found that mutants lacking dsk or its receptors are hypersensitive to noxious heat. Through a combination of genetic, behavioral, histological, and Ca2+ imaging analyses, we subsequently revealed neurons involved in DSK-mediated nociceptive regulation at a single-cell resolution and identified a DSKergic descending neuronal pathway that inhibits nociception. This study provides the first evidence for a descending modulatory mechanism of nociception from the brain in a non-mammalian species that is mediated by the evolutionarily conserved CCK system, raising the possibility that the descending inhibition is an ancient mechanism to regulate nociception.

Introduction

Minimizing tissue damage is a fundamental task for all animals to increase their chance of survival. Thus, elucidating the principles of nociception, the neural process detecting and encoding potentially tissue-damaging stimuli, is critical to understanding the molecular and neural mechanisms implementing adaptive behaviors and their evolution. Nociceptors are sensory neurons specialized to detect harmful stimuli, whose activation triggers downstream nociceptive circuits and nocifensive responses (1). Since the activities of nociceptors and downstream nociceptive circuits are tightly linked to pain perception in humans, unveiling the mechanisms of nociception is also crucial to a better understanding of human pain mechanisms (2, 3).

Descending inhibition has been suggested to be a pivotal mechanism in the modulation of nociception and pain in mammals. Since the discovery that electrical stimulations of parts of the midbrain in rats enabled surgical operations without anesthetics (4), mammalian descending nociceptive pathways have been implicated in various analgesic phenomena/treatments and the development of chronic pain states, suggesting their critical role in modulating nociception and pain (5, 6). However, brain-mediated modulatory mechanisms of nociception such as descending inhibition have currently been identified only in mammals, despite a high degree of commonality across species in the peripheral nociceptive mechanisms (7, 8). Therefore, it is unknown whether the descending modulation is a de novo mechanism typical of the highly developed mammalian central nervous system (CNS) or a conserved control also present in simpler animals.

The descending nociceptive-modulatory systems in mammals have been revealed to involve various neurochemical pathways (5, 6, 9); among these, the cholecystokinin (CCK) system is one of the most extensively characterized (6, 9). In rodents, CCK signaling plays a crucial role in facilitating nociception by counteracting the opioidergic systems in the periaqueductal gray (PAG)—rostral ventral medulla (RVM)—spinal descending pathway (10–13) and inhibiting nociception through the central amygdala (CeA)—PAG—spinal pathway (14). In humans, CCK signaling has been implicated in nocebo hyperalgesia, mediated by the descending nociceptive control system (15). The CCK system is very well-conserved and has been implicated in several common physiological functions among bilaterian species (16–20). However, whether CCK is functionally involved in regulating nociception outside of mammals remains unknown.

Drosulfakinin (DSK), a neuropeptide homologous to CCK, was identified in the fruit fly Drosophila melanogaster (19). Fly DSK is reportedly involved in modulating many physiological functions shared with mammals, including gut functions, anxiety, aggression, memory, feeding, synaptic functions, and courtship behaviors (19, 21–23). After the discovery of stereotyped nociceptive escape behavior called rolling and polymodal Class IV md (C4da) nociceptors (24, 25), the larval Drosophila has been successfully utilized to identify evolutionarily conserved and previously uncharacterized molecular pathways in nociception (26–34) and circuitry mechanisms in the ventral nerve cord (VNC; the invertebrate equivalent of the spinal cord) to compute multimodal sensory stimuli and select nociceptive escape strategies (35–40). Previous studies have demonstrated that neuropeptidergic systems also participate in regulating nociception in Drosophila (39, 41–44). However, the role of fly DSK in nociception remains elusive.

Here, using a collective approach of genetic, behavioral, histological, and Ca2+ imaging analyses, we pursued the mechanisms of DSK-mediated nociceptive regulation and demonstrate that the DSK system constitutes a descending inhibitory pathway of nociception from the brain to the VNC in larval Drosophila.

Results

DSK signaling negatively regulates thermal nociception

Through a thermal nociception screen using the nocifensive rolling response of Drosophila larvae, we found that a deletion mutant line of the dsk gene showed thermal hypersensitivity with a significantly shorter latency in their response to a 42 °C probe than the controls, suggesting that DSK plays a role in negatively regulating nociception (Figure 1A and B). A genomic fragment containing the wild-type dsk gene, and no other neuropeptide genes, significantly rescued the thermal hypersensitivity of the dsk mutants (Figure 1B), confirming that dsk is responsible for the thermal hypersensitivity.

DSK signaling is involved in negatively regulating nociception.

(A) Predicted amino acid sequences of pro-DSK peptide in the wild-type (top) and dsksk1 mutants (bottom). Due to a 5-base deletion at the +94-99 position in the coding sequence, dsksk1mutants are predicted to produce a largely truncated pro-DSK peptide unable to be processed to active DSK peptides. Red letters represent residues different from the wild-type and asterisks indicate stop codons. (B) Hypersensitivity of dsksk1 mutants to a 42°C thermal probe. Significantly shortened latencies of dsksk1 mutants (n = 143) compared with the controls (n = 164) were recovered in genomic rescue animals (Dp(3;1)2-2; dsksk1, n = 139). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. (C) Predicted amino acid sequence of CCKLR-17D1 in the wild-type and CCKLR-17D1 mutants. CCKLR-17D1Δ1, CCKLR-17D1Δ2, and CCKLR-17D1Δ3 have 17-base (+1573-1589), 2-base (+1575-1576) and 32-base (+1573-1604) deletions in the coding sequence respectively, which result in large frameshifts completely abolishing the sixth and seventh transmembrane domains (TMs). (D) CCKLR-17D1Δ1(n = 54), CCKLR-17D1Δ2 (n = 51) and CCKLR-17D1Δ3mutants (n = 57) all showed significantly shorter latencies than control (n = 38). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. (E) Predicted amino acid sequence of CCKLR-17D3 in the wild-type and CCKLR-17D3 mutants. CCKLR-17D3Δ1, CCKLR-17D3Δ2, and CCKLR-17D3Δ3 possess 8-base (+474-481), 32-base (+457-488), and 5-base (+472-476) deletions in the coding sequence respectively, which result in large frameshifts and truncation in the middle of the second TM. (F) CCKLR-17D3Δ1 (n = 32), CCKLR-17D3Δ2 (n = 38), and CCKLR-17D3Δ3mutants (n = 36) all exhibited significantly shorter latencies than control (n = 70). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points.

DSK has been shown to activate two G-protein coupled receptors, CCKLR-17D1 and CCKLR-17D3 (45, 46), which are orthologous to the mammalian CCK receptors, CCKAR (also known as CCK1) and CCKBR (also known as CCK2) (16–18). To test whether these receptors mediate DSK signaling in nociception, we generated deletion mutants for CCKLR-17D1 and CCKLR-17D3 using CRISPR/Cas9 genome editing and tested them for thermal nociception (Figure 1C-F). When stimulated with a 42 °C probe, three independent deletion lines of the CCKLR-17D1 and CCKLR-17D3 exhibited thermal hypersensitivity (Figure 1D and F), further supporting the role of DSK signaling in the negative regulation of thermal nociception. To examine whether the phenotypes of CCKLR-17D1 and CCKLR-17D3 mutants are classified as hyperalgesia (hypersensitivity to normally noxious stimuli) or allodynia (abnormal hypersensitivity to normally innocuous stimuli), we tested the receptor mutants with a 38 °C probe, which is close to the threshold of larval thermal nociception (39 °C) (24). We found that the responses of the DSK receptor mutants were indistinguishable from the controls, indicating that the DSK receptor mutants are hypersensitive to suprathreshold thermal stimuli, thus hyperalgesic (Figure 1-figure supplement 1).

We have noticed that the yw control strain, which was used by us to generate the dsk and receptor deletion mutants, showed relatively longer response latencies to the 42 °C probe compared to the other control strains used in this study. This may be attributed to the effect of the genetic background, although, presently, the cause for this difference is unknown.

Two groups of brain neurons expressing DSK are responsible for regulating nociception

We attempted to identify DSK-expressing cells in the larval CNS that are responsible for regulating nociception. Unlike mammalian CCK, which is expressed in the CNS and gastrointestinal system, Drosophila DSK is expressed in the CNS but not in the gut (19, 47–50). Previous immunohistochemical studies have reported putative DSK-expressing cells in the larval CNS (47, 49). However, since the specificity of the DSK antibodies has not been validated with null mutants in the former studies, it has been a concern that the reported DSK-expressing cells could include non-DSK cells that express the other neuropeptides sharing the C-terminal RFa motif with DSK (51). Consistent with this concern, we found that an antibody against crustacean FLRFa, an FMRFa-like neuropeptide with the C-terminal RFa motif, gives rise to a comparable staining pattern to that of the previously reported DSK antibodies, visualizing cells designated as insulin-producing cells (IPCs; referred to as SP3 in Nichols and Lim, (1996) (47)), MP1, SP1, SP2, Sv, SE2, Tv1-3, T2dm, and A8 in the larval CNS (Figure 2A) (47, 49).

DSK is expressed in two groups of larval brain neurons.

(A) Representative image of DSK-GAL4 expression and anti-FLRFa staining in the wild-type larval CNS. (B) Representative image of anti-FLRFa staining in dsksk1 larval CNS. (C) Representative images of DSK-GAL4 expression and anti-FLRFa staining in the control larval brains. Arrows and arrowheads indicate MP1 and Sv neurons, respectively. Co-expressions of DSK-GAL4 and anti-FLRFa were observed in MP1 and Sv in all of the examined samples (n = 27/27), but in IPCs only in 7% (n = 2/27). (D) Representative images of DSK-GAL4 expression and anti-FLRFa staining in the dsksk1 larval brains. Arrows and arrowheads indicate MP1 and Sv neurons, respectively. (E) Representative images of DSK-GAL4 expression and anti-FLRFa staining in the larval brains of the genomic rescue genotype (Dp(3;1)2-2/Y; +/lexA-rCD2::RFP UAS-mCD8::GFP; dsksk1 DSK-GAL4/dsksk1). Arrows and arrowheads indicate MP1 and Sv neurons, respectively. (F) An example image showing the expression of DSK-2A-GAL4, a 2A-GAL4 knock-in line of the dsk gene, more faithfully recapitulating its endogenous expression. Arrows and arrowheads indicate MP1 and Sv neurons, respectively. MP1 neurons were labeled in 100% of DSK-2AGAL4 samples (n = 41/41), while a single IPC in 51.2% (n = 21/41), multiple IPCs in 24.4% (n = 10/41), and Sv neurons in 17.1% (n = 7/41) of examined samples. All scale bars represent 20 µm.

To identify bona fide DSK-expressing cells responsible for regulating nociception, we performed anti-FLRFa staining in multiple dsk null alleles and found that the staining signals persist in all but two pairs of neurons, MP1 and Sv, in the mutant CNS (Figure 2B-D and Figure 2-figure supplement 1A-D), suggesting that these two pairs of neurons are the only neurons in the larval CNS that express DSK. The genomic rescue fragment that rescued the thermal hypersensitivity of dsk mutants (Figure 1B) also restored the anti-FLRFa signals in MP1 and Sv in a dsk mutant background (Figure 2E), suggesting that the dsk gene is responsible for the anti-FLRFa signals in these neurons as well as the thermal nociceptive responses of larvae. This expression pattern was further corroborated by the transgenic reporter line DSK-GAL4 and the 2A-GAL4 knock-in reporter line DSK-2A-GAL4 (52), both of which we found were expressed only in MP1 and Sv neurons among the anti-FLRFa-positive neurons (Figure 2A, C, and F; Figure 2-figure supplement 1B-F). We also found that anti-FLRFa, DSK-GAL4, and DSK-2A-GAL4 do not visualize larval peripheral neurons including C4da nociceptors (Figure 2A and F; Figure 2-figure supplement 1G). Taken together, these results identified two sets of brain neurons, MP1 and Sv, as the DSK-expressing cells that are involved in regulating nociception in larvae.

CCKLR-17D1 in Goro neurons functions to negatively regulate nociception

We sought the potential target cells of DSK signaling for regulating nociception. Since DSK receptor mutants were thermally hypersensitive consistently to dsk mutants (Figure 1B, D, and F), neurons in the larval nociceptive circuit that express DSK receptors were promising candidates. To visualize the cells expressing DSK receptors, we generated T2A-GAL4 knock-ins in CCKLR genes. Both CCKLR-17D1-T2A-GAL4 and CCKLR-17D3-T2A-GAL4 were widely expressed in the larval CNS, predominantly in neuronal cells (Figure 3A and B; Figure 3-figure supplement 1A-D). By performing double-labeling experiments, we found that CCKLR-17D1-T2A-GAL4 and CCKLR-17D3-T2A-GAL4 are not expressed in the nociceptors at the periphery (Figure 3C and D; Figure 3-figure supplement 1E and F), or in nociceptive interneurons including Basin1-4 (53), A08n (36, 41), and DnB neurons (35) in the larval VNC (Figure 3-figure supplement 1G-L). However, we found that they are expressed in Goro neurons (Figure 3E and F), which are the fourth-order nociceptive interneurons located in the larval VNC (53).

DSK receptors are expressed in Goro neurons in the larval VNC.

(A and B) Representative images of CCKLR-17D1-T2A-GAL4 (A) and CCKLR-17D3-T2A-GAL4 (B) expressions in the larval CNS. (C and D) Representative images showing double-labeling of CCKLR-17D1-T2A-GAL4 (C) and CCKLR-17D3-T2A-GAL4 (D) with C4da nociceptors (R38A10-lexA). Arrows and arrowheads indicate cell bodies of C4da nociceptors and es cells, respectively. (E and F) Representative images showing double-labeling of CCKLR-17D1-T2A-GAL4 (E) and CCKLR-17D3-T2A-GAL4 (F) with Goro neurons (arrowheads, R69E06-lexA). Expression patterns were confirmed in multiple samples (n = 7 and 5). All scale bars represent 20 µm.

To test whether the DSK receptors in Goro neurons are functionally important for regulating nociception, we performed RNAi and rescue experiments using R69E06-GAL4 that marks Goro neurons (53). RNAi knockdown of CCKLR-17D1, but not CCKLR-17D3, in R69E06-GAL4 neurons induced thermal hypersensitivity (Figure 4A and Figure 4-figure supplement 1A). R69E06-GAL4 is also expressed in multiple neurons in the larval brain other than Goro neurons in the VNC (53). However, the thermal hypersensitivity was not observed when the expression of CCKLR-17D1 RNAi was excluded from Goro neurons by tsh-GAL80, pointing to the requirement of CCKLR-17D1 in Goro neurons (Figure 4-figure supplement 1B-D). Consistent with these RNAi results, expressing wild-type CCKLR-17D1, but not CCKLR-17D3, with R69E06-GAL4 rescued the hypersensitivity of the respective mutants (Figure 4B and C). Furthermore, the morphology of Goro neurons was not affected by CCKLR-17D1 knockdown (Figure 4-figure supplement 1E). Therefore, we conclude that CCKLR-17D1 functions in Goro neurons to negatively regulate thermal nociception, but the function of CCKLR-17D3 in nociceptive regulations resides elsewhere.

CCKLR-17D1 in Goro neurons is necessary and sufficient for normal thermal nociception.

(A) RNAi of CCKLR-17D1 and CCKLR-17D3 using a Goro-GAL4 line R69E06-GAL4. Expressing CCKLR-17D1 RNAi (R69E06-GAL4 x yv; JF02644, n = 66) but not CCKLR-17D3 RNAi (R69E06-GAL4 x yv; JF02968, n = 95) with R69E06-GAL4 caused significantly shorter latencies to 42 °C than controls (R69E06-GAL4 x yv; attp2, n = 46). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001, NS (non-significant) p > 0.05 Steel’s test. (B) Rescue of CCKLR-17D1 with R69E06-GAL4. The shortened latencies of CCKLR-17D1Δ1 mutants observed in UAS controls (CCKLR-17D1Δ1; UAS-CCKLR-17D1/+, n = 61) or GAL4 controls (CCKLR-17D1Δ1; R69E06-GAL4/+, n = 42) were restored in the rescue genotype (CCKLR-17D1Δ1; UAS-CCKLR-17D1/+; R69E06-GAL4/+, n = 52). (Left) Histograms (Right) Box plots of latencies. *** p < 0.001 Steel’s test. (C) Rescue of CCKLR-17D3 with R69E06-GAL4. The shortened latencies of CCKLR-17D3Δ1mutants were unaltered in the rescue genotype (CCKLR-17D3Δ1; UAS-CCKLR-17D3/+; R69E06-GAL4/+, n = 130) compared with UAS controls (CCKLR-17D3Δ1; UAS-CCKLR-17D3/+, n = 40) or GAL4 controls (CCKLR-17D3Δ1; R69E06-GAL4/+, n = 68). (Left) Histograms (Right) Box plots of latencies. NS (non-significant) p > 0.05 Steel’s test. All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points.

Activation of Goro neurons elicits nocifensive rolling in larvae (53). Hence, if CCKLR-17D1 functions in Goro neurons to negatively regulate nociception, the lack of CCKLR-17D1 should sensitize these neurons to noxious heat. We directly addressed this hypothesis by using a Ca2+ imaging technique we had previously developed (32, 35), whereby we monitored GCaMP6m signals in Goro neurons while locally applying thermal ramp stimuli to the larval body wall. In our thermal ramp stimulation protocol, the temperature around the larval CNS hardly reached 28 °C during the heat ramp stimulations (27.0 ± 0.2 °C at the peak, n = 12), suggesting that the larval nociceptive system in the CNS that functions as an internal sensor for fast temperature increase (54) was unlikely to be activated.

Goro neurons in CCKLR-17D1Δ1 mutants showed a significantly steeper GCaMP6m signal increase from 40 to 50 °C in comparison with the wild-type controls (Figure 5A and B, Video 1 and 2). Since the baseline fluorescence levels of GCaMP6m were not significantly different in the wild-type and CCKLR-17D1Δ1 mutants (Figure 5C), these data demonstrate that Goro neurons in CCKLR-17D1Δ1 mutants are specifically sensitized to a noxious range of heat. Suppressing CCKLR-17D1 by RNAi in Goro neurons also induced significantly sensitized responses of Goro to noxious temperatures of 44–49 °C (Figure 5D-F, Video 3 and 4). In contrast, Goro neurons in CCKLR-17D3Δ1mutants exhibited GCaMP6m signals that were mildly elevated but largely parallel compared with that of the controls (Figure 5-figure supplement 1, Video 5 and 6), providing further evidence for the major functioning of CCKLR-17D3 in nociceptive regulation outside Goro neurons. Considered together, these data demonstrate that CCKLR-17D1 functions to negatively regulate the activity of Goro neurons, thereby attenuating behavioral nociceptive responses.

Goro neurons lacking CCKLR-17D1 show sensitized responses to noxious heat.

(A) Representative still images showing thermal activation of Goro neurons in the controls (top, yw/Y; R69E06-GAL4 UAS-GCaMP6m/+) and CCKLR-17D1Δ1 mutants (bottom, CCKLR-17D1Δ1/Y; R69E06-GAL4 UAS-GCaMP6m/+). See also Video 1 and 2. (B) Average percent increase of GCaMP6m fluorescence intensity relative to the baseline (ΔF/F0) during heat ramp stimulations in controls (n = 15) and CCKLR-17D1Δ1 (n = 15). * p < 0.05, ** p < 0.01 Mann-Whitney’s U-test. (C) Basal GCaMP6m signal levels (F0) in controls (n = 15) and CCKLR-17D1Δ1 mutants (n = 15). p > 0.6 Mann-Whitney’s U-test. (D) Representative stills showing thermal activation of the controls (top, R69E06-GAL4 UAS-GCaMP6m x yv; attp2) and Goro neurons expressing CCKLR-17D1 RNAi (bottom, R69E06-GAL4 UAS-GCaMP6m x yv; JF02644). See also Video 3 and 4. (E) Average percent increase of GCaMP6m fluorescence intensity relative to the baseline (ΔF/F0) during heat ramp stimulations in controls (n = 15) and CCKLR-17D1 RNAi (n = 16). * p < 0.05 Mann-Whitney’s U-test. (F) Basal GCaMP6m levels (F0) in controls (n = 15) and CCKLR-17D1 RNAi (n = 16). p > 0.6 Mann–Whitney’s U-test. Error bars represent standard error. All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points. All scale bars represent 20 µm.

Neuronal projections and thermal responsiveness of DSK-expressing neurons

If CCKLR-17D1 is involved in regulating the activity of Goro neurons, how can DSK be conveyed from the brain to the VNC? We noticed that some of the DSK-GAL4 positive brain neurons sent descending neural processes to the VNC (Figure 6A), which were all anti-FLRFa positive (Figure 6B). Furthermore, the anti-FLRFa signals in the descending projections were completely absent in the dsk mutants (Figure 6C), suggesting that these descending projections likely originated from the DSK-expressing brain cells, namely MP1 and/or Sv. In analyzing DSK-2A-GAL4, we found that MP1 neurons in fact sent the descending projections to the VNC (Figure 6D). To further understand the projection patterns of MP1 and Sv neurons in detail, we performed single-cell labeling using an FLP-out technique and revealed that MP1 neurons sent descending projections contralaterally to the VNC (Figure 6E), while Sv neurons projected within the brain (Figure 6F). In comparison with the longitudinal processes to the VNC, MP1 neural processes in the brain possessed few anti-FLRFa positive puncta, which represent DSK in these neurons (Figure 6G). Furthermore, when the somatodendritic marker UAS-Denmark (55) and the synaptic vesicle marker UAS-syt::eGFP were expressed in MP1 neurons, Denmark preferentially localized in the neural processes within the brain while syt::eGFP strongly accumulated in the processes descending to the VNC (Figure 6H). Collectively, these data demonstrate that MP1 neurons project DSK-positive descending axons to the VNC from the brain.

Projection patterns of DSK-expressing neurons in the larval CNS.

(A) Representative image showing descending axons from DSK-GAL4 positive brain neurons (green) to the larval VNC. (B) Representative image showing all descending projections labeled by DSK-GAL4 (green) harbor punctate anti-FLRFa signals (magenta) in the wild-type. (C) Representative image showing the complete absence of punctate anti-FLRFa signals (magenta) in the descending projections labeled by DSK-GAL4 (green) in the dsksk1 mutants. (D) Image showing the descending projections of MP1 neurons (arrows) marked by DSK-2A-GAL4. (E) Image showing a single FLP-out clone of MP1 neuron (arrow) contralaterally sending descending axons to the larval VNC. The yellow dashed line indicates the midline. (F) Image showing two FLP-out clones of Sv neurons (arrowheads) sending ascending axons to both sides of the brain. (G) A projection image showing the MP1 neurites in the brain. Note that few anti-FLRFa-positve puncta are associated with MP1 neurites within the brain. Similar expression patterns were observed in multiple samples (n = 3). The arrow indicates the anti-FLRFa-positve MP1 soma. (H) Representative image showing localizations of syt::eGFP (green) and Denmark (magenta) in MP1 neurons (DSK-2A-GAL4 x UAS-syt::eGFP UAS-Denmark). Similar expression patterns were observed in multiple samples (n = 7). All scale bars represent 20 µm.

Next, we investigated whether some of these DSK-expressing neurons are responsive to noxious thermal stimuli. In order to examine the responsiveness of multiple neurons in intact larvae, we took the snapshot approach using CaMPARI, a genetically-encoded Ca2+ sensor whose fluorescence irreversibly changes from the green to the red region of the spectrum in response to irradiation with 405 nm UV light in a Ca2+ dependent manner, and is thus useful to monitor the activities of multiple neurons simultaneously in intact animals (56). When the larvae expressing CaMPARI2 (an improved version of CaMPARI (57)) in Goro neurons were stimulated by local thermal ramp stimuli under a 405 nm UV light, a significantly increased CaMPARI2 photoconversion was detected compared to the control group that had only been irradiated with 405nm UV (Figure 7A and B), confirming that by using CaMPARI2 in our experimental setup we managed to successfully capture the increased activity of Goro neurons in response to noxious heat in intact larvae. We then tested the responsiveness of DSK-2A-GAL4 neurons using the same stimulation and photoconversion protocol and found that none of the MP1 or Sv neurons, nor IPCs showed increased CaMPARI2 photoconversion in response to local thermal ramp stimuli (Figure 7A, C-E). Interestingly, MP1 neurons exhibited high CaMPARI2 photoconversion regardless of the presence of local heat ramp stimuli, which was even comparable to that observed in Goro neurons activated by noxious heat stimuli (Figure 7A and C). These data suggest that none of the DSK-2A-GAL4 neurons are clearly responsive to noxious heat, but the markedly high CaMPARI2 photoconversion observed in MP1 neurons raises the possibility that these neurons may be active regardless of the presence of noxious heat stimuli.

Activity and thermal responsiveness of DSK-expressing neurons.

(A) Box plots of the quantified ratio of CaMPARI2 green signals and photoconverted CaMPARI2 red signals detected by anti-CaMPARI-red in Goro, MP1, Sv, and IPCs (Red/Green) with or without local heat ramp stimulation. Goro neurons showed a significantly higher Red/Green ratio of CaMPARI2 when local heat ramp stimulation was applied to the larvae (n = 16 and 20, p < 0.001, Mann-Whitney’s U-test). MP1 neurons showed a high Red/Green ratio of CaMPARI2 both with and without local heat stimulation (n = 21 and 20, p > 0.3, Mann-Whitney’s U-test). The Red/Green ratio of CaMPARI2 was low in Sv compared to MP1 neurons, and CaMPARI2 photoconversion was not increased by local heat stimulation (n = 9 and 10, p > 0.5, Mann-Whitney’s U-test). IPCs also exhibited low CaMPARI2 photoconversion regardless of the presence of local heat ramp stimulation (n = 17 and 16, p > 0.25, Mann-Whitney’s U-test). (B-E) Representative images showing CaMPARI2 green signals (green) and anti-CaMPARI-red signals (magenta) in Goro (B), MP1 (C), Sv (D), and IPCs (E). All box plots show median (middle line) and 25th to 75th percentiles, with whiskers indicating the smallest to the largest data points. Scale bars represent 10 µm.

Connectivity between DSKergic and Goro neurons

To gain more insights into the connectivity between DSK-expressing brain neurons and Goro neurons, we further examined the anatomical relationship between MP1 axons and Goro neurons in the VNC. Performing double-labeling experiments, we found that MP1 axons with anti-FLRFa signals (representing DSK in MP1 axons) partially overlapped with Goro neurites in the thoracic segments of the larval VNC (Figure 8A). GFP reconstitution was reproducibly detected between MP1 and Goro neurons using the CD4-GRASP (GFP Reconstitution Across Synaptic Partners) system (58, 59), thus confirming that MP1 axons and Goro neurons are indeed in close proximity (Figure 8B). Since the CD4-GRASP is known to detect general cell-cell contacts (60, 61), we further investigated whether MP1 and Goro neurons are synaptically connected by using the nSyb-GRASP technique, which specifically enables synapse detection by localizing one of the split-GFP fragments in presynaptic sites (62). However, no signals of reconstituted GFP were detected between MP1 and Goro with nSyb-GRASP, although we found proximate localizations of nSyb-GFP1-10 in the MP1 axons with the Goro neurites expressing CD4::GFP11 (Figure 8C). We also employed the trans-Tango system (63) as another tool to detect synaptic connectivity. However, when the trans-Tango ligand was expressed by DSK-2A-GAL4, synaptic connectivity between MP1 and Goro was not indicated since the postsynaptic activation of trans-Tango signals was not observed in Goro neurons, although some trans-Tango-positive neurons were observed around Goro neurons (Figure 8D).

Synaptic connectivity between MP1 axons and Goro neurons is not detected.

(A) Representative images of the larval VNC showing partially overlapping MP1 axons (green) and processes of Goro neurons (blue) that are associated with anti-FLRFa signals (red). MP1 axons partially overlapping Goro neurons were similarly observed in all examined samples (n = 5). (B) Representative image showing CD4-GRASP experiments between MP1 axons and Goro neurons (magenta). Signals of reconstituted GFP (GRASP, green) were detected in all examined samples (n = 5). (C) Representative image showing nSyb-GRASP experiments between MP1 axons (blue) and Goro neurons (red). Signals of reconstituted GFP (GRASP, green) were not detectable in any of the examined samples (n = 5). (D) Representative image showing trans-Tango experiments using DSK-2A-GAL4 (R69E06-lexA, lexAop-rCD2::RFP UAS-mCD8::GFP; DSK-2A-GAL4 x UAS-myrGFP, QUAS-mtdTomato::3xHA; trans-Tango). Transsynaptic activation of Tango signals (blue; mtdTomato::3xHA detected by anti-HA) was undetectable in Goro neurons (red; rCD2::RFP detected by anti-CD2). Green represents MP1 axons labeled by DSK-2A-GAL4 (myrGFP and mCD8::GFP detected by anti-GFP). Signals of mtdTom::3xHA in Goro neurons were not detected in any of the examined samples (n = 4). Arrows indicate Goro somata. All scale bars represent 20 µm, except for the insets in (A) (5 µm).

Although we could not find positive evidence for synaptic connectivity between MP1 axons and Goro neurons, we further tested the hypothesis that DSK-expressing neurons in the brain may have functional connectivity to Goro neurons. To achieve this, we first attempted artificial activation experiments of DSK-2A-GAL4 neurons using UAS-dTRPA1 to determine whether activating DSK-expressing neurons actually attenuates larval nociception. dTRPA1 is a cation channel gated by warm temperature (> 29 °C) that has been used as a tool to artificially activate neurons of interest in Drosophila (64). When the larvae expressing dTRPA1 in C4da nociceptors with ppk1.9-GAL4 were placed in a chamber at 35°C, dTRPA1-induced thermogenetic activation of C4da nociceptors caused nocifensive rolling responses within two seconds in the majority of animals, as reported previously (65) (Figure 9A and B). In contrast, when the larvae expressing dTRPA1 simultaneously in ppk1.9-GAL4 and DSK-2A-GAL4 were placed in a chamber at 35°C, the additional activation of DSK-2A-GAL4 neurons to C4da nociceptors resulted in a markedly reduced percentage of larvae showing rolling responses within two seconds and in significantly lengthened latencies to exhibit a rolling response (Figure 9A and B). None of the animals expressing dTRPA1 with DSK-2A-GAL4 alone showed a rolling response within 10 seconds (Figure 9A).

Thermogenetic activation of DSK-2A-GAL4 neurons inhibits larval nociception and Goro neurons.

(A and B) Thermogenetic activations of nociceptors and/or DSK-2A-GAL4 neurons using UAS-dTRPA1. (A) Histogram representing the percentage of larvae exhibiting nociceptive rolling. Upon thermogenetic activations, 55% of larvae expressing dTRPA1 in nociceptors (ppk-GAL4 > dTRPA1; ppk1.9-GAL4 x UAS-dTRPA1, n = 58) exhibited nociceptive rolling within 2 seconds. In contrast, only 11% of larvae with simultaneous activations of nociceptors and DSK-2A-GAL4 showed rolling within 2 seconds (ppk-GAL4 & DSK-2A-GAL4 > dTRPA1; ppk1.9-GAL4 DSK-2A-GAL4 x UAS-dTRPA1, n = 54). Thermogenetic activations of DSK-2A-GAL4 neurons alone (DSK-2A-GAL4 > dTRPA1; DSK-2A-GAL4 x UAS-dTRPA1, n = 40) did not trigger nociceptive responses even after 10 seconds. (B) Box plots showing latencies. Thermogenetic activation of DSK-2A-GAL4 simultaneously with nociceptors caused a significantly longer latency to rolling responses (Mann-Whitney U-test, *** p < 0.001). (C and D) CaMPARI2 snapshot imaging of the activity of Goro neurons activated by thermogenetic stimulation of C4da nociceptors. (C) Box plots of the ratio of CaMPARI2 green signals and photoconverted CaMPARI2 red signals detected by anti-CaMPARI-red in Goro neurons (Red/Green). Compared to the larvae with thermogenetic activation of C4da nociceptors only (ppk-GAL4 > dTRPA1; ppk1.9-GAL4, UAS-dTRPA1 x R69E06-lexA; lexAop-CaMPARI2; n = 17), the larvae with simultaneous activation of C4da and DSK-2A-GAL4 neurons (ppk-GAL4 & DSK-2A-GAL4 > dTRPA1; ppk1.9-GAL4, UAS-dTRPA1; DSK-2A-GAL4 x R69E06-lexA; lexAop-CaMPARI2; n = 19) showed significantly decreased CaMPARI2 photoconversion in Goro neurons (Mann-Whitney U-test, * p < 0.025). (D) Representative images showing CaMPARI2 green signals (green) and anti-CaMPARI-red signals (magenta) in Goro neurons of animals with thermogenetic C4da activation (ppk-GAL4 > dTRPA1) and with simultaneous thermogenetic activation of C4da and DSK-2A-GAL4 neurons (ppk-GAL4 & DSK-2A-GAL4 > dTRPA1). All box plots show median (middle line) and 25th to 75th percentiles with whiskers indicating the smallest to the largest data points. Scale bars represent 10 µm.

The above data suggest that activation of DSK-2A-GAL4 neurons inhibits larval nociception. We then investigated whether activating DSK-2A-GAL4 neurons could attenuate the activity of Goro neurons, by combining the artificial activation experiments described above with CaMPARI2 snapshot imaging of Goro neurons. Indeed, CaMPARI2 photoconversion in Goro neurons induced by the thermogenetic activations of C4da nociceptors was significantly decreased by 35% with the simultaneous activation of DSK-2A-GAL4 neurons (Figure 9C and D). Given that MP1 neurons were the only cells that were labeled by DSK-2A-GAL4 with 100% reproducibility, while the expression of DSK-2A-GAL4 in IPCs and Sv neurons was minor and stochastic (Figure 2F), the observed inhibition of behavioral nociception and Goro neurons caused by activating DSK-2A-GAL4 neurons is mostly attributable to the activation of MP1 neurons. Thus, these data support the functional connectivity between DSKergic MP1 neurons in the brain and Goro neurons.

Discussion

In this study, we have demonstrated that (1) DSK and its receptors CCKLR-17D1 and CCKLR-17D3 are involved in negatively regulating thermal nociception, (2) Two sets of brain neurons, MP1 and Sv, express DSK in the larval nervous system, (3) One of the DSK receptors (CCKLR-17D1) functions in Goro neurons in the VNC to negatively regulate thermal nociception, and (4) Thermogenetic activation of DSK-2A-GAL4 neurons attenuates both larval nociception and the activity of Goro neurons. Based on these results, we propose that the DSKergic neurons regulating the activity of Goro neurons constitute a descending inhibitory pathway of nociception from the brain to the VNC in larval Drosophila (Figure 10). To our knowledge, our findings represent the first evidence of a descending mechanism modulating nociception from the brain in a non-mammalian species.

A schematic model of the DSKergic descending inhibitory pathway of nociception in larval Drosophila.

DSK-expressing MP1 and/or Sv neurons in the brain secrete DSK peptides. DSK in the larval VNC activates the CCKLR-17D1 receptor expressed in Goro neurons, which subsequently inhibits the activity of Goro neurons, and ultimately larval nociceptive rolling responses.

DSKergic descending system as a physiological modulator of nociception

DSK has been implicated in multiple physiological and developmental processes in Drosophila (19, 21–23). Previous studies have shown that dsk and CCKLR-17D1 mutants exhibit significant reductions of synaptic growth and excitability in larval neuromuscular junctions (NMJ) and larval locomotion under bright light (45, 66), suggesting the importance of the DSK/CCKLR-17D1 signaling pathway in the developmental processes of motoneurons. However, in this study, no major developmental defects were observed in Goro neurons with CCKLR-17D1 RNAi (Figure 4-figure supplement 1E). Furthermore, the simultaneous thermogenetic activations of DSK-2A-GAL4 neurons and C4da nociceptors inhibited larval nociception and the activity of Goro neurons (Figure 9). Given the pronociceptive role of Goro neurons, their reduced synaptic or neuronal activity should cause nociceptive insensitivity or reduced Ca2+ responses, which is contradicting to our observation that CCKLR-17D1 knockdown in Goro produced the thermal hypersensitivity and exaggerated Ca2+ responses (Figure 4 and 5). Thus, these data consistently support a physiological role for DSKergic descending signals in the modulation of the activity of Goro neurons rather than a developmental role, highlighting the functional differences of the DSK/CCKLR-17D1 pathway between the NMJ and the nociceptive system in the CNS.

DSK in larval IPCs dispensable for nociceptive modulation

A previous study has reported the expression of DSK in a population of larval IPCs and its functions in responding to starving conditions (49). However, we observed that anti-FLRFa signals in IPCs persisted in dsk null mutants and that anti-FLRFa signals in IPCs hardly overlap with either DSK-GAL4 or DSK-2A-GAL4 expressions (Figure 2 and Figure 2-figure supplement 1). Although we used a different antibody from that in the study mentioned above, the number of IPCs visualized by anti-FLRFa was comparable to that of cells visualized by anti-DSK (Figure 2-figure supplement 1) (49). Regarding the functions of IPCs in nociception, Im et al. showed that silencing dilp2-GAL4 positive IPCs has no significant effect on the baseline nociceptive responses (67). In our CaMPARI2 imaging, IPCs marked by DSK-2A-GAL4 showed very low baseline activity and no responsiveness to noxious heat (Figure 7). Thus, the current and previous studies strongly suggest that IPCs are likely irrelevant for DSK-mediated nociceptive regulation, at least under normal conditions, although it is still possible for DSK to be expressed stochastically in a limited population of IPCs.

Functional mechanisms of DSKergic nociceptive inhibitory system

Although our data suggest the existence of a DSKergic descending inhibitory system from the brain to the Goro neurons in the VNC, there are multiple remaining questions for future studies on the mechanisms by which DSK-expressing neurons in the brain regulate Goro neurons: First, whether both MP1 and Sv neurons are involved in nociceptive inhibition remains unclear. Our findings consistently point to MP1 neurons over Sv neurons as a major source of DSK regulating Goro neurons. However, given the potential of DSK for distant actions through diffusion, the possibility of Sv neurons communicating with Goro neurons cannot be eliminated. Further analysis using finer genetic/neuronal tools will be necessary to tease the functions of MP1 and Sv neurons apart and understand the functional mechanism of the DSKergic system in the regulation of larval nociception.

Second, how this DSKergic system is activated to regulate nociception in vivo needs further investigation. We observed very high CaMPARI2 photoconversion in MP1 neurons without noxious heat stimuli, which did not further increase in response to noxious heat application (Figure 7). These data raise the hypothesis that MP1 neurons may function as a tonically active brake for nociceptive circuits, rather than a simple negative feedback system that is activated by nociceptive stimuli as input. The tonically active model of MP1 function is also consistent with our data showing that dsk and CCLR-17D1 mutations, as well as CCLR-17D1 RNAi in Goro neurons, all caused thermal hypersensitivity under normal conditions (Figure 1 and 4). Up to this point, no external or internal signals have been directly shown to be the input into larval DSKergic neurons. However, since DSK has been implicated in the regulation of several physiological processes such as hunger and stress, sensory or molecular signals that are involved in these physiological processes have the potential to serve as upstream cues that activate DSKergic neurons. Alternatively, it is also possible that MP1 neurons may have spontaneous activity. Further physiological studies on the activity and responsiveness of DSKergic neurons would be required to elucidate how the DSKergic system functions to regulate nociception in larvae under normal conditions.

Third, the transmission mechanisms of DSK from the brain to Goro neurons need further clarification. Our nSyb-GRASP and trans-Tango experiments consistently showed negative results, indicating no synaptic connectivity between MP1 descending axons and Goro neurons (Figure 8). Thus, the interaction between DSK from DSKergic neurons in the brain and Goro neurons in the VNC may be mediated non-synaptically through volume transmission, as described in many neuropeptidergic systems (68). Many of the CCKLR-expressing neurons in the VNC are distantly located from the descending MP1 axons (Figure 3 and 6). Since it is unlikely that all these CCKLR-expressing neurons are synaptically connected to MP1 axons, neuronal communications through volume transmission can be fairly assumed for DSKergic systems in the larval VNC. More detailed analyses at an electron microscopic level of the circuitry connectivity between MP1 and Goro neurons and of the localization of DSK-containing vesicles in DSKergic neurons as well as DSK receptors in Goro neurons would be necessary to further clarify the transmission mechanisms of DSK from the brain to Goro neurons.

Multiple molecular pathways for DSK signaling in the regulation of nociception

The data presented in this study also suggest that DSK signaling could regulate larval nociception through multiple pathways other than the DSK/CCKLR-17D1 system. For example, while CCKLR-17D3 mutants exhibited as severe thermal hypersensitivity as CCKLR-17D1 mutants (Figure 1D and F), our RNAi and rescue experiments failed to locate the function of CCKLR-17D3 to Goro neurons (Figure 4A and C). Our GCaMP Ca2+ imaging also revealed that Goro neurons lacking CCKLR-17D3 showed modestly sensitized responses to noxious heat (Figure 5-figure supplement 1). These results suggest the major functioning of CCKLR-17D3 in Goro-independent nociceptive pathways. We also observed that DSK receptor mutants exhibited more severe thermal phenotypes than dsk mutants (Figure 1), which may indicate the presence of unidentified DSK- or CCKLR-dependent signaling pathways in nociception-related cells. Further research is apparently necessary to reveal the whole picture of nociceptive regulations mediated by DSK and its receptors in larval Drosophila.

Potential conservation of CCK-mediated descending nociceptive controls

The CCK system is thought to be one of the most ancient neuropeptide systems, suggested to have multiple common physiological functions across taxa (16–20). In this study, we demonstrate that CCKergic signaling in Drosophila participates in nociceptive modulation through a descending inhibitory pathway similarly to the mammalian CCK system, adding new evidence about the conserved physiological roles of CCK.

Unlike the peripheral nociceptive systems, it is still challenging to align the Drosophila and mammalian CCKergic descending pathways due to low homologies of the CNS structures between Drosophila and mammals, wide-spread CCK expression in the mammalian CNS, and the multiple roles of the CCKergic systems in mammalian nociceptive controls (10–14, 69–73). However, the common usage of an orthologous molecular pathway in descending controls of nociception between the two evolutionarily distant clades raises a fascinating new hypothesis that the descending control from the brain may also be an ancient, conserved mechanism of nociception, which has emerged in the common ancestor of protostomes and deuterostomes. It will be of interest for future research to investigate whether the role of CCK signaling in descending nociceptive controls is also present in other species.

Potentials of non-mammalian models to study the mechanisms of pain modulation and pain pathology in the CNS

Non-mammalian model systems have been increasingly recognized as powerful tools to identify novel pain-related molecular pathways (2, 74–77). However, the utilization of these models has so far been mostly limited to the research on peripheral pain pathophysiology, and few studies have used them to investigate central pain pathophysiology (2, 78). Descending nociceptive control mechanisms are crucial for central pain modulation and have been implicated in the development of chronic pain states in humans (5, 6). Thus, the current study opens the door to a new approach to using powerful neurogenetic tools and the simpler nervous system of Drosophila for elucidating the functional principles of descending nociceptive systems, which may potentially contribute to our understanding of the mechanisms underlying central pain modulation and pain pathology due to dysfunctions of descending modulatory pathways.

Supporting information